---
title: "Download AquaMatch Data"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sept 14, 2026"
---

**Pipeline step 00 · R · run from `regional_clarity/` · chunks are `eval = FALSE`; run by hand only when AquaMatch changes upstream**

- **Purpose:** Download the AquaMatch Secchi disc depth (SDD) and siteSR data, stack siteSR across Landsat missions, and apply the Gardner handoff coefficients referenced to Landsat 7.
- **Inputs:** AquaMatch via `AquaMatchr` (needs `EDI_API_key` in `.Renviron`).
- **Outputs:** `aquamatch_files/downloads/`: `sdd_cache.parquet`, `siteSR_DSWE1_LS7corr.parquet`, `sceneMetadata_Landsat*.csv`, site metadata and handoff CSVs.
- **Next:** [01 · Filter AquaMatch data](01_filter_AquaMatch_Data.Rmd)

# Background

This step downloads the AquaSat SDD and siteSR data and applies the Gardner-style
handoff coefficients for use in a modeling workflow to estimate SDD across the 
intermountain west. Only run this script if there have been upstream changes
to the AquaMatch dataset, since downloading files takes some time to complete.

Knitting this file will not result in any download, as all chunks are set to
`eval = FALSE`.

## Setup

Load the packages used in this step. `AquaMatchr` handles all communication
with the AquaMatch data packages on EDI; `arrow` and `data.table` handle the
large parquet and feather files it produces.

```{r libraries, eval=FALSE}
# load libraries
library(AquaMatchr) # if install needed: remotes::install_github(repo = "AquaSat/AquaMatchr")
library(data.table)
library(arrow)
library(tidyverse)
```

Authenticate with the Environmental Data Initiative (EDI) repository before
requesting any files. `AquaMatchr` can fail to prompt for credentials on its
own, so this logs in explicitly using the `EDI_API_key` stored in `.Renviron`.

```{r edi-login, eval=FALSE}
# login to EDI - this is a work around for error popping up in AquaMatchr
# requires EDIutils, which is required by AquaMatchr so no need to import
EDIutils::login(key = Sys.getenv('EDI_API_key'))
```

# Download AquaMatch data

Set a download location for the files. Everything downloaded or built in
this step is written to `aquamatch_files/downloads/`, which is the input
directory for [01 · Filter AquaMatch data](01_filter_AquaMatch_Data.Rmd).

```{r aquamatch-dir, eval=FALSE}
# note location for file storage when downloading via AquaMatchr
aquamatch_dir <- "aquamatch_files/downloads"
dir.create(aquamatch_dir, recursive = T)
```

Use `download_parameters()` to retrieve the newest version of the AquaMatch
Secchi disc depth (SDD) dataset, and cache it locally as a parquet file so
later steps can read it without re-downloading.

```{r download-data, eval=FALSE}
# use AquaMatchr for data gathering
sdd <- download_parameters(parameters = "sdd", 
                           version = "newest")
# get it out of a list
sdd <- sdd[["sdd"]]
write_dataset(dataset = sdd, 
              path = file.path(aquamatch_dir, "sdd_cache.parquet"), 
              format = "parquet")
rm(sdd)
```

Next, download the siteSR data: the Landsat surface reflectance summaries
extracted at each AquaMatch site. These arrive as separate files per Landsat
mission. The algal mask is not applied here.

```{r download-siteSR, eval=FALSE}
download_siteSR(save_location = aquamatch_dir, 
                version = "newest", 
                algal_mask = FALSE)
```

Stack the per-mission siteSR files into a single dataset so that all Landsat
missions share one table, saved as a feather file.

```{r siteSR-stack, eval=FALSE}
build_sr(which_sr = "siteSR",
         sr_location = aquamatch_dir,
         algal_mask = FALSE,
         save = TRUE, 
         save_location = file.path(aquamatch_dir, "siteSR_DSWE1_full_concatenation.feather")
)

```

Atmospheric correction and differing sensor specifications across Landsat missions
can create artificial patterns in the surface reflectance product, so each
mission is harmonized to a common reference before modeling. This chunk
converts the stacked feather file to parquet, then applies the Gardner-style
polynomial handoff coefficients that AquaMatch provides, using Landsat 7 as
the reference sensor. The result, `siteSR_DSWE1_LS7corr.parquet`, is the
reflectance input to the rest of the pipeline.

```{r apply-handsoff, eval=FALSE}
# workaround for feather/parquet issue
open_dataset(file.path(aquamatch_dir, "siteSR_DSWE1_full_concatenation.feather"), 
             format = "feather") %>% 
  write_dataset(dataset = .,  
                path = file.path(aquamatch_dir, "siteSR_DSWE1_full_concatenation.parquet"),
                format = "parquet")

# and apply handoffs
apply_handoffs(
  input_path = file.path(aquamatch_dir, "siteSR_DSWE1_full_concatenation.parquet"),
  handoff_path = list.files(aquamatch_dir, full.names = T)[grepl("handoff", list.files(aquamatch_dir))],
  correction_method = "Gardner_poly",
  sat_target = "LS7",
  algal_mask = "FALSE",
  save_location = "aquamatch_files/downloads/siteSR_DSWE1_LS7corr.parquet"
)
```

Finally, download the scene-level metadata for each Landsat mission (e.g.,
cloud cover, sun angle, and acquisition time). [01](01_filter_AquaMatch_Data.Rmd)
uses these fields to filter the reflectance data.

```{r sitesr-scene, eval=FALSE}
download_sceneMetadata(save_location = aquamatch_dir, product = "siteSR")
```

