---
title: "Filter AquaMatch Data"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 19, 2026"
output:
  html_document
---

**Pipeline step 01 · R · run from the repository root**

- **Purpose:** Define the 6-state HUC4 area of interest, then quality-filter SDD (tier/flag QC, per-site RANSAC) and siteSR (scene metadata QA, majority-vote band RANSAC) within it.
- **Inputs:** `aquamatch_files/downloads/`: `sdd_cache.parquet`, `siteSR_DSWE1_LS7corr.parquet`, `sceneMetadata_Landsat*.csv`, site metadata CSV; national WBD geodatabase.
- **Outputs:** `aquamatch_files/`: `sdd_quality_filter.parquet`, `siteSR_DSWE1_regional_band_filtered.parquet`, `siteSR_DSWE1_regional_band_ransac_diagnostics.parquet`, `huc4_filtered.gdb`, `states.gdb`, `row_loss_log_01.csv`.
- **Previous:** [00 · Download AquaMatch data](00_download_AquaMatch_data.Rmd) · **Next:** [02a · Make matchups](02a_make_matches.Rmd)

This step applies some filters to the SDD and siteSR data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

Load packages and set the two working directories: the raw downloads from
[00](00_download_AquaMatch_data.Rmd) and the destination for this step's
filtered outputs. `tigris` supplies state boundaries, `hydrogeofetch` the
HUC4 watershed boundaries, and `igraph` the site de-duplication further
below.

```{r libraries, , message = FALSE, warning=FALSE}
# load libraries
library(igraph)
library(parallel)
library(arrow)
library(sf)
library(tigris)
library(hydrogeofetch)
library(tidyverse)

aquamatch_raw_data_dir <- "regional_clarity/aquamatch_files/downloads"
filtered_aquamatch_dir <- "regional_clarity/aquamatch_files"
```

## Regional AOI

Every filter below (SDD quality/RANSAC filtering, siteSR outlier filtering)
operates per-site and needs no broader/national context, so we determine our
6-state region's HUC4 footprint here, up front, and scope both pipelines down
to it before any of the expensive per-site processing runs - rather than
waiting until `02a_make_matches.Rmd` to subset down to it after matching and
after the costly RANSAC fits have already run on out-of-region data.

First, pull the boundaries of the six states that make up the region
(Colorado, Wyoming, Utah, Idaho, Montana, and New Mexico) and dissolve them
into a single polygon.

```{r states-aoi, message = FALSE, warning=FALSE}
states <- states(progress_bar = FALSE) %>%
  filter(STUSPS %in% c("CO", "WY", "UT", "ID", "MT", "NM"))
states_union <- states %>% st_union()
```

Grab all the HUC4's touching our 6-state AOI. We prefer a live NHD/WBD
query via `hydrogeofetch::get_huc()`; that service has intermittently timed
out against the USGS pygeoapi backend in the past (even retried
state-by-state), so we cap it at 30 seconds and fall back to the WBD
national geodatabase already cached locally (`aquamatch_files/wbd_national`)
if it errors or doesn't return in time.

```{r huc4}
huc4_cache_gdb <- file.path(filtered_aquamatch_dir,
                            "wbd_national/WBD_National_GDB/WBD_National_GDB.gdb")

HUC4_national <- tryCatch(
  # run get_huc() in a subprocess (not a signal-based timeout in this
  # session) - interrupting a GDAL/curl call mid-flight in-process can leave
  # the driver in a bad state for the fallback st_read() below; a subprocess
  # can just be killed cleanly on timeout instead
  callr::r(
    function(aoi) {
      library(hydrogeofetch)
      get_huc(AOI = aoi, type = "huc04")
    },
    args = list(aoi = states_union),
    timeout = 30
  ),
  error = function(e) {
    message("Live HUC4 fetch failed or timed out (", conditionMessage(e),
            "); falling back to the cached WBD national geodatabase.")
    st_read(huc4_cache_gdb, layer = "WBDHU4", quiet = TRUE) %>%
      # WBD ships with Z values and a compound CRS (NAD83 + NAVD88 height);
      # transforming a 2D CRS into that compound CRS silently breaks (no
      # vertical grid to invent a Z), so drop Z here - horizontal datum is
      # then plain NAD83
      st_zm(drop = TRUE, what = "ZM") %>%
      rename_with(tolower)
  }
)

# standardize the geometry column name to "geometry" (FileGDB layers are
# often read in with the geometry column named "shape"; get_huc() already
# names it "geometry")
geom_col <- attr(HUC4_national, "sf_column")
if (geom_col != "geometry") {
  names(HUC4_national)[names(HUC4_national) == geom_col] <- "geometry"
  st_geometry(HUC4_national) <- "geometry"
}

# reduce from the national/live layer to just the HUC4's touching our
# 6-state AOI
HUC4v <- HUC4_national %>%
  st_transform(st_crs(states_union)) %>%
  st_make_valid() %>%
  st_filter(states_union)
```

And then assess overlapping area - here we want to include HUCs that overlap
the 6-state area by at least 50%.

```{r huc4-overlap}
overlap_areas <- st_make_valid(HUC4v) %>%
  st_intersection(st_make_valid(states_union)) %>%
  mutate(overlap_area = st_area(geometry)) %>%
  st_drop_geometry() %>%
  select(huc4, overlap_area)

# calculate proportion overlap, filter for those greater than/equal to 50%
HUC4_filtered <- HUC4v %>%
  left_join(overlap_areas, by = "huc4") %>%
  mutate(
    overlap_area = coalesce(overlap_area, units::set_units(0, "m^2")),
    overlap_frac = as.numeric(overlap_area / st_area(geometry))
  ) %>%
  filter(overlap_frac >= 0.50)
```

Save the HUC4 and state boundaries as their own geodatabases so this and
later scripts can just read these back in instead of re-running the HUC4
download/filter and `states()` pull.

```{r export-boundaries}
st_write(HUC4_filtered,
         dsn = file.path(filtered_aquamatch_dir, "huc4_filtered.gdb"),
         layer = "huc4_filtered",
         driver = "OpenFileGDB",
         append = FALSE,
         quiet = TRUE)

st_write(states,
         dsn = file.path(filtered_aquamatch_dir, "states.gdb"),
         layer = "states",
         driver = "OpenFileGDB",
         append = FALSE,
         quiet = TRUE)
```

Initialize a row-count log to track how much data survives each filtering
step below, so we can summarize retention/loss once the full pipeline
(through `02a_make_matches.Rmd`) has run.

```{r init-row-log}
row_log <- tibble(pipeline = character(), step = character(),
                  n_rows = double(), n_sites = double())

log_step <- function(log, pipeline, step, n_rows, n_sites = NA_real_) {
  bind_rows(log, tibble(pipeline = pipeline, step = step,
                        n_rows = n_rows, n_sites = n_sites))
}
```

## SDD data filtering

This section applies a ransac-style filter to sites with greater numbers of
SDD observations to remove obvious outliers based on historical patterns and
trends.

The chunk below runs in three stages:

1. **Quality filters.** Keep only higher-tier AquaMatch records (`tier < 3`)
   with a plausible depth (≥ 0.1 m), no miscellaneous or bottom-hit flags,
   agreement among any aggregated same-day measurements (CV ≤ 0.25), and a
   monitoring location typed as a lake, reservoir, or impoundment.
2. **Regional restriction.** Keep only sites that fall inside the HUC4
   footprint defined above.
3. **Per-site RANSAC.** `ransac_site()` repeatedly fits a linear trend
   through time to a random subsample of 10 observations at a site and
   counts how often each observation lands within a robust (MAD-based)
   threshold of that fit. Observations that are inliers in at least 60% of
   the iterations are retained. An observation is never flagged if it lies
   within 1 m of the fitted trend, or if it falls inside the range of the
   site's own inliers (narrowed at each end by 15% of the site's mean SDD). 
   Sites with 10 or fewer observations are left unlabeled (`NA`).
   Same-day aggregated values that already passed the CV check are always
   kept. The labeled table is cached to `sdd_quality_filter.parquet`.

The row-count log records the number of rows and sites remaining after each stage.

```{r}
sdd_raw_path <- file.path(aquamatch_raw_data_dir, "sdd_cache.parquet")

sdd <- open_dataset(sdd_raw_path)

sdd_filtered_path <- file.path(filtered_aquamatch_dir, "sdd_quality_filter.parquet")

row_log <- log_step(row_log, "SDD", "0_raw_nationwide",
                    n_rows = sdd %>% summarize(n = n()) %>% collect() %>% pull(n))

sdd_qc <- sdd %>%
  filter(
    # do some SDD filters - must be in higher tiers of data
    tier < 3,
    # must have a reasonable  value
    harmonized_value >= 0.1,
    # if there are aggregated rows, they generally agree
    harmonized_value_cv <= 0.25 | is.na(harmonized_value_cv),
    # no flags
    misc_flag == 0,
    # not a bottom hit
    depth_flag == 0,
    # and we'll just grab lake/res/impoundment data
    ResolvedMonitoringLocationTypeName == "Lake, Reservoir, Impoundment"
  ) %>%
  collect()

row_log <- log_step(row_log, "SDD", "1_quality_tier_flag_filters",
                    n_rows = nrow(sdd_qc),
                    n_sites = n_distinct(sdd_qc$MonitoringLocationIdentifier))

# restrict to sites within our 6-state region's HUC4 footprint (see
# "Regional AOI" above) - RANSAC and everything else here runs per-site, so
# there's no reason to spend time fitting sites this project will never use
sdd_site_pts <- sdd_qc %>%
  distinct(MonitoringLocationIdentifier, lat, lon) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  st_transform(st_crs(HUC4_filtered)) %>%
  st_filter(HUC4_filtered)

sdd_qc <- sdd_qc %>%
  filter(MonitoringLocationIdentifier %in% sdd_site_pts$MonitoringLocationIdentifier)

row_log <- log_step(row_log, "SDD", "2_regional_HUC4_restriction",
                    n_rows = nrow(sdd_qc),
                    n_sites = n_distinct(sdd_qc$MonitoringLocationIdentifier))

# run a ransac-like algorithm per site to remove outliers
ransac_site <- function(date, one_site_sdd,
                        n_iter = 50,
                        mad_thresh = 2,
                        n_sd = 3,
                        range_buffer_frac = 0.15,
                        min_diff = 1,
                        min_sample = 10,
                        vote_frac = 0.6) {
  n <- length(one_site_sdd)
  if (n < min_sample + 1) return(rep(NA, n))  # too few points to run ransac

  t <- as.numeric(date)
  # tally, per point, how many valid iterations called it an inlier
  inlier_votes <- rep(0L, n)
  n_valid_iter <- 0L

  for (i in 1:n_iter) {
    idx <- sample(n, min_sample)
    if (length(unique(t[idx])) < 2) next  # avoid degenerate fit (same date)
    out_idx <- setdiff(seq_len(n), idx)

    fit <- lm(one_site_sdd[idx] ~ t[idx])
    pred <- coef(fit)[1] + coef(fit)[2] * t
    resid <- one_site_sdd - pred
    # scale should come from held-out points only - points used to fit the
    # line have near-zero residual by construction, so including them
    # deflates the scale (and threshold) whenever min_sample is close to n,
    # making a single held-out point look like an outlier over a trivial
    # difference. use MAD rather than sd() here: sd() is itself inflated by
    # a cluster of genuine outliers landing in the held-out set (a handful of
    # large same-direction residuals drag the mean and blow up the squared
    # deviations), which loosens the threshold and lets that same cluster
    # vote itself in as inliers. MAD is robust to that as long as the
    # cluster is a minority of the held-out set, since it centers on the
    # median rather than the mean.
    resid_scale <- if (length(out_idx) >= 2) {
      mad(resid[out_idx], constant = mad_thresh, na.rm = TRUE)
    } else {
      # too few held-out points (n == min_sample + 1) to trust their own MAD -
      # fall back to all-points MAD (deflated, but not degenerate) instead
      mad(resid, constant = mad_thresh, na.rm = TRUE)
    }
    # never flag a point closer than min_diff to the fit line, regardless of
    # how tight the scale-based threshold is - a < 1 m disagreement isn't
    # meaningfully an outlier for Secchi depth.
    threshold <- max(n_sd * resid_scale, min_diff, na.rm = TRUE)
    inliers <- !is.na(resid) & abs(resid) < threshold

    inlier_votes <- inlier_votes + as.integer(inliers)
    n_valid_iter <- n_valid_iter + 1L
  }

  # indicate when no iteration ever produced a usable fit
  if (n_valid_iter == 0) return(rep(NA, n))

  # a point is an inlier if at least vote_frac of the valid iterations agreed
  best_inliers <- (inlier_votes / n_valid_iter) >= vote_frac

  # a flagged point can't be implausible for this site if its value falls
  # within the range other accepted (inlier) observations already span here -
  # deviating from the local trend line isn't grounds to flag it when the
  # site itself has already produced values in that range. this matters most
  # for low-n sites, where a short fit sample leaves little room to
  # distinguish a real trend from a value the site has simply shown before
  if (any(best_inliers)) {
    inlier_range <- range(one_site_sdd[best_inliers], na.rm = TRUE)
    # scale the buffer to this site's typical SDD rather than using a flat
    # meters value - 0.5 m is a huge margin at a 1 m-visibility site and
    # negligible at a 10 m-visibility site
    range_buffer <- range_buffer_frac * mean(one_site_sdd[best_inliers], na.rm = TRUE)
    best_inliers <- best_inliers | (one_site_sdd >= inlier_range[1] + range_buffer &
                                      one_site_sdd <= inlier_range[2] - range_buffer)
  }

  best_inliers
}


if (!file.exists(sdd_filtered_path)) {
  set.seed(43)
  sdd_qc %>%
    arrange(harmonized_local_time)%>%
    mutate(is_inlier = ransac_site(harmonized_local_time, harmonized_value),
           .by = MonitoringLocationIdentifier) %>%
    # also reverse the flag in cases where multiple measurements are made
    # since we've already confirmed multiple measurements agree (value_cv <= 0.25)
    mutate(is_inlier = if_else(harmonized_row_count > 1 &
                                 (!is_inlier | is.na(is_inlier)),
                               TRUE,
                               is_inlier)) %>%
    write_parquet(sdd_filtered_path)
}
```

Check whether flagging concentrates at the SDD extremes rather than spreading
evenly across the distribution. A high-end concentration here isn't necessarily
a problem: we expect a lot of what lands in
the far right tail of raw SDD is genuinely implausible (deeper than the
water body, deeper than the instrument's rope, transcription errors, etc.)
rather than real high-clarity variability - the point of this check is to
tell those two cases apart before trusting the filter.

```{r ransac-check-sdd-tails}
sdd_ransac_check <- read_parquet(sdd_filtered_path)
```

Let's look at highly-flagged sites as a spot-check. Each plot is one site's
SDD time series, with points colored by whether RANSAC kept them (four or
more flagged observations).

```{r}
high_flag_sites <- sdd_ransac_check %>%
  summarize(n = n(),
            n_flagged = sum(is_inlier == FALSE),
            mean_sdd = mean(harmonized_value),
            .by = c(MonitoringLocationIdentifier, lat, lon)) %>%
  arrange(-n_flagged) %>%
  filter(n_flagged >= 4)

map(.x = high_flag_sites$MonitoringLocationIdentifier,
    ~ sdd_ransac_check %>%
      filter(MonitoringLocationIdentifier == .x) %>%
      ggplot(., aes(x = ActivityStartDate,
                    y = harmonized_value,
                    color = is_inlier)) +
      geom_point() +
      labs(title = .x))
```

There are clearly a couple overflagging examples here, but it's a bit out
of scope to refine the ransac further.

Record the number of SDD rows that continue to matching. Observations flagged as
outliers (`is_inlier == FALSE`) are excluded. Unresolved observations (`NA`,
from sites with too few points for RANSAC) are kept.

```{r}
# rows that will actually reach 02a_make_matches.Rmd - is_inlier == FALSE is
# excluded there before matching; NA (RANSAC never resolved, e.g. too few
# points) is kept, same treatment as the siteSR band RANSAC below
sdd_for_matching <- sdd_ransac_check %>%
  filter(is_inlier | is.na(is_inlier))

row_log <- log_step(row_log, "SDD", "3_ransac_inliers_for_matching",
                    n_rows = nrow(sdd_for_matching),
                    n_sites = n_distinct(sdd_for_matching$MonitoringLocationIdentifier))
```

Re-summarize and look again, this time lowering the bar to sites with three
or more flagged observations to see a broader set of examples:

```{r}
high_flag_sites <- sdd_ransac_check %>%
  summarize(n = n(),
            n_flagged = sum(is_inlier == FALSE),
            mean_sdd = mean(harmonized_value),
            .by = c(MonitoringLocationIdentifier, lat, lon)) %>%
  arrange(-n_flagged) %>%
  filter(n_flagged >= 3)

map(.x = high_flag_sites$MonitoringLocationIdentifier,
    ~ sdd_ransac_check %>%
      filter(MonitoringLocationIdentifier == .x) %>%
      ggplot(., aes(x = ActivityStartDate,
                    y = harmonized_value,
                    color = is_inlier)) +
      geom_point() +
      labs(title = .x))
```

Let's also look to see what fraction of observations are indicated to be
outliers across the range of observed Secchi depth. Each bar shows the
share of observations in that depth bin that were flagged (red) versus
kept (grey):

```{r}
ggplot(sdd_ransac_check, aes(x = harmonized_value, fill = is_inlier)) +
  geom_histogram(position = "fill", bins = 30) +
  scale_fill_manual(values = c(`TRUE` = "grey70", `FALSE` = "red")) +
  labs(x = "observed Secchi (m)", y = "fraction of observations",
       fill = "inlier") +
  theme_bw()
```

This all looks okay.

## Remote Sensing Filters

Scene-level metadata (cloud cover, image quality, geometric accuracy, etc.)
isn't carried through siteSR's per-site aggregation, so we download it
separately (via `00_download_AquaMatch_data.Rmd`) and join it on below via
`sat_id`.

```{r download-scene-metadata, message = FALSE, warning=FALSE}
scene_metadata_paths <- list.files(
  aquamatch_raw_data_dir,
  pattern = "sceneMetadata_Landsat(457|89)\\.csv$",
  full.names = TRUE
)

if (length(scene_metadata_paths) == 0) {
  stop("No scene-level metadata found in ", aquamatch_raw_data_dir,
       " - run 00_download_AquaMatch_data.Rmd first.")
}

# LS457 and LS89 have slightly different columns (e.g. a single
# IMAGE_QUALITY vs. split IMAGE_QUALITY_OLI/_TIRS), so bind_rows() unions
# them rather than assuming identical schemas
scene_metadata <- map(.x = scene_metadata_paths,
                      .f = ~ read_csv(., show_col_types = FALSE) %>%
                        filter(CLOUD_COVER < 20)) %>%
  bind_rows()
```

Next, build the site table used to scope the reflectance data. Starting from
AquaMatch's site metadata, keep only sites whose point buffer
intersects a single waterbody (`number_int_wb == 1`), carry the waterbody
identifiers and shoreline flags through, and restrict to sites inside the
regional HUC4 footprint. Each site is tagged with its `huc4`.

```{r}
sites_geom_qa <- read_csv(file.path(aquamatch_raw_data_dir, "siteSR_collated_WQP_NWIS_sites_with_NHD_info_2025-06-04.csv"),
                  show_col_types = FALSE) %>%
  # make sure the point extent only touches one waterbody
  filter(number_int_wb == 1) %>%
  select(siteSR_id, wb_nhd_id, wb_gnis_name, wb_areasqkm,
         flag_wb, flag_optical_shoreline,
         flag_thermal_TM_shoreline, flag_thermal_ETM_shoreline,
         flag_thermal_TIRS_shoreline, WGS84_Latitude, WGS84_Longitude)

# restrict to sites within our 6-state region's HUC4 footprint (see
# "Regional AOI" above) - none of the site-level QA below (de-duping,
# ransac) needs regional context, so scope down before it runs rather than
# waiting until 02a_make_matches.Rmd to subset down after matching
sites <- sites_geom_qa %>%
  st_as_sf(coords = c("WGS84_Longitude", "WGS84_Latitude"), crs = 4326,
           remove = FALSE) %>%
  st_transform(st_crs(HUC4_filtered)) %>%
  st_join(HUC4_filtered %>% select(huc4), left = FALSE) %>%
  # guard against a site sitting exactly on a shared HUC4 boundary, which
  # could otherwise match more than one polygon in the spatial join above
  distinct(siteSR_id, .keep_all = TRUE) %>%
  st_drop_geometry()
```

Different `siteSR_id` values (different source prefixes - `WQP_*`, `AM_*`,
`NWIS_*`, etc.) can still be the exact same physical monitoring location,
or two locations close enough together that Landsat can't tell them
apart. Two sites within 100m of each other but assigned to *different*
waterbodies aren't the same location - that's more likely two real,
distinct water features close together (or an NHD assignment error) than
one site measured twice - so sites are only collapsed together when
they're both within 100m of each other *and* share the same `wb_nhd_id`.
Grouping is transitive (connected components over that adjacency), so a
chain of near-duplicates collapses into a single `location_id` even if
its two most-distant members are themselves more than 100m apart.

Note that we do *not* filter out shoreline-flagged sites here (`flag_wb`,
`flag_optical_shoreline`, and the thermal shoreline flags are carried
through, not filtered on) - shoreline proximity turns out to be a useful
predictive feature rather than pure contamination, so a later script
collapses these into a `shore_flag` feature instead of discarding the data.

```{r site-location-dedup}
site_pts <- sites %>%
  st_as_sf(coords = c("WGS84_Longitude", "WGS84_Latitude"), crs = 4326,
           remove = FALSE)

# sparse adjacency: which sites (including itself) fall within 100m of
# each site
site_within_100m <- st_is_within_distance(site_pts, dist = 100)

# ...then restrict those pairs to ones sharing a waterbody - equality is
# symmetric, so this keeps the adjacency symmetric (if j survives in i's
# filtered neighbor list, i survives in j's) for the connected-components
# step below
wb_nhd_id <- sites$wb_nhd_id
site_within_100m_same_wb <- Map(
  function(nbrs, wb) nbrs[wb_nhd_id[nbrs] == wb],
  site_within_100m, wb_nhd_id
)

# connected components over that adjacency - this is what makes the
# grouping transitive rather than strictly pairwise
site_components <- components(
  graph_from_adj_list(site_within_100m_same_wb, mode = "all")
)$membership

pad <- nchar(as.character(max(site_components)))
sites <- sites %>%
  mutate(location_id = sprintf(paste0("loc_%0", pad, "d"), site_components))

# one row per de-duplicated location, listing the individual siteSR_id
# values collapsed into it
site_location_groups <- sites %>%
  summarize(n_sites = n(),
            siteSR_ids = list(siteSR_id),
            WGS84_Latitude = mean(WGS84_Latitude),
            WGS84_Longitude = mean(WGS84_Longitude),
            .by = location_id) %>%
  arrange(-n_sites)

print(paste0(nrow(site_location_groups %>% filter(n_sites > 1)), " locations have
             more than one site in the 100m area in the same waterbody"))

site_location_groups %>%
  filter(n_sites > 20)

# for locations with more than one siteSR_id within 100m (see
# site_location_groups above), keep just the first siteSR_id per
# location_id, so a duplicated physical site doesn't feed the filter below
# more than once
representative_sites <- sites %>%
  distinct(location_id, .keep_all = TRUE) %>%
  pull(siteSR_id)

```


## Filter reflectance outliers before matching

AquaMatch's siteSR data is 167.5 million rows across 342,704 sites nationwide, we
know that there are still outliers in the reflectance data, so we filter
reflectance outliers here, *before* `match_siteSR_to_WQP()` ever sees the data,
rather than after matching. We want to make sure
outlier filtering happens before matches are made so that a flagged bad RS
data point doesn't remove the matched row down with it.

Instead of running on the full nationwide siteSR dataset, we use
five scoping/QA steps to keep the expensive per-site filter below from
processing sites this project will never use or trusting scenes/sites it
shouldn't:

- **HUC2 pre-filter**: a cheap, coarse pushdown restriction to the HUC2
basins that overlap this project's 6-state regional AOI, ahead of the
precise (but not pushdown-able) HUC4/geometry scoping below.
- **Site geometry and regional HUC4 QA**: `siteSR_id` is restricted to the
`sites` table built above, which drops sites whose buffer overlaps with more
than one waterbody (`number_int_wb == 1`) *and* restricts to sites falling
within our 6-state region's HUC4 footprint (see "Regional AOI" at the top of
this script) - this is the precise regional scoping that used to happen only
at the very end of `02a_make_matches.Rmd`, after matching.
- **Metadata QA**: `pCount_dswe1/pCount_dswe_gt0 > 0.5`, `pCount_dswe1 > 8`,
`prop_clouds == 0` pretty generous filters to remove possbile contamination
- **Scene-level cloud QA**: `sat_id` is further restricted to scenes in
`scene_metadata` with
`CLOUD_COVER < 20`, catching a partly-cloudy scene that could have aerosol impacts
on the temp and optical data
- **Site de-duping**: The `location_id` dedup built alongside `sites` above
(collapsing same-waterbody sites within 100m) is used to identify one
representative site to process through the ransac filter

Corrected bands, not raw bands, feed the ransac filter - reflectance isn't
comparable across missions without harmonizing first. The cross-sensor
harmonization (Gardner et al. coefficients, relative to Landsat 7) was
already applied once in `00_download_AquaMatch_data.Rmd` via
`apply_handoffs()`, so we read those corrected columns
(`*_corr_7`) straight from its output (`siteSR_DSWE1_LS7corr.parquet`)
rather than recomputing them here.

The filter itself (`ransac_site_multiband()`) is the same per-site,
random-subsample RANSAC-like approach as `ransac_site()` above, just fitting
all seven corrected channels (six reflectance bands plus surface
temperature) per iteration instead of one, standardized by each channel's
own MAD-based scale from a single all-data fit: reflectance is a shakier fit
against a straight line over a site's full record than SDD is (blooms,
turbidity, and seasonality move bands around much more), so instead of a raw
residual threshold in reflectance units we standardize each band's
per-iteration residual by its own MAD and combine into one z-score per
point, then threshold on that combined score.

The regression inside the iteration loop is written as closed-form OLS
rather than `lm()` to reduce processing time and we parallellize to further
optimize the time need here.

```{r siteSR-regional-band-ransac-setup}
siteSR_regional_filtered_path <- file.path(filtered_aquamatch_dir, "siteSR_DSWE1_regional_band_filtered.parquet")

# a second, diagnostics-only cache that retains all rows and inlier labels
# so later figures render on every knit without re-running the RANSAC fit
siteSR_regional_corr_diag_path <- file.path(filtered_aquamatch_dir, "siteSR_DSWE1_regional_band_ransac_diagnostics.parquet")

# store the ransac funciton
#
# Majority-vote across iterations (vote_frac >= cutoff), the same design as
# ransac_site() above - not winner-take-all (picking whichever single random
# iteration produced the most inliers), which this function used until it
# was redesigned to match: winner-take-all is far more sensitive to one
# lucky/unlucky random draw, and with only a 5-day match window and
# Landsat's sparse revisit, aggressively flagging scenes on that kind of
# noisy basis can zero out the *only* available candidate image for a field
# sample rather than just degrading match quality.
ransac_site_multiband <- function(date, band_mat,
                                  n_iter = 150,
                                  n_mad = 2.5,
                                  # per-band sensor-noise floor
                                  min_diff = c(red_corr_7 = 0.01, green_corr_7 = 0.01,
                                               blue_corr_7 = 0.01, nir_corr_7 = 0.01,
                                               swir1_corr_7 = 0.01, swir2_corr_7 = 0.01,
                                               surfacetemp_corr_7 = 1),
                                  min_sample = 10,
                                  vote_frac_cutoff = 0.5) {
  n <- nrow(band_mat)
  nb <- ncol(band_mat)
  if (n < min_sample + 1) return(rep(NA, n))  # too few points to run ransac

  # match by name rather than assuming column order lines up with min_diff
  min_diff_b <- min_diff[colnames(band_mat)]

  t <- as.numeric(date)

  # fixed per-band residual scale from a single all-data fit (MAD is robust
  # up to ~50% contamination, well above what we expect here) - reused for
  # every candidate subsample below instead of recomputed per iteration
  band_scale <- vapply(seq_len(nb), function(b) {
    y <- band_mat[, b]
    cf <- ols_fit(t, y)
    pred <- cf["intercept"] + cf["slope"] * t
    resid <- y - pred
    s <- mad(resid, na.rm = TRUE)
    if (is.na(s) || s == 0) s <- sd(resid, na.rm = TRUE)
    s
  }, numeric(1))

  # tally, per point, how many valid iterations called it an inlier
  inlier_votes <- rep(0L, n)
  n_valid_iter <- 0L

  for (i in 1:n_iter) {
    idx <- sample(n, min_sample)
    if (length(unique(t[idx])) < 2) next  # avoid degenerate fit (same date)

    # fit each band vs time on the *same* subsample, standardize each band's
    # residuals by its fixed scale, then combine - keeps one band's scale
    # from dominating and keeps the threshold comparable across candidate
    # lines of very different quality
    z <- matrix(NA_real_, nrow = n, ncol = nb)
    for (b in seq_len(nb)) {
      y <- band_mat[, b]
      if (length(unique(y[idx])) < 2 || is.na(band_scale[b]) || band_scale[b] == 0) next

      cf <- ols_fit(t[idx], y[idx])
      if (anyNA(cf)) next
      pred <- cf["intercept"] + cf["slope"] * t
      resid <- y - pred
      z[, b] <- abs(resid) / band_scale[b]
      # a raw difference this small is within sensor noise for this band -
      # never let it count as outlier evidence, regardless of how tight
      # band_scale is
      z[abs(resid) < min_diff_b[b], b] <- 0
    }

    combined_z <- rowMeans(z, na.rm = TRUE)
    inliers <- !is.na(combined_z) & combined_z < n_mad

    inlier_votes <- inlier_votes + as.integer(inliers)
    n_valid_iter <- n_valid_iter + 1L
  }

  # no iteration ever produced a usable fit - we haven't determined
  # anything about this site, so say so rather than leaving every point
  # flagged by the rep(0L, n) initialization
  if (n_valid_iter == 0) return(rep(NA, n))

  # a point is an inlier if at least vote_frac_cutoff of the valid
  # iterations agreed
  best_inliers <- (inlier_votes / n_valid_iter) >= vote_frac_cutoff

  # a flagged point can't be implausible for this site if every one of its
  # band values falls within the range other accepted (inlier) points
  # already span for that band - mirrors the same rescue in ransac_site()
  if (any(best_inliers)) {
    in_range <- vapply(seq_len(nb), function(b) {
      y <- band_mat[, b]
      accepted <- y[best_inliers]
      if (length(accepted) == 0 || all(is.na(accepted))) return(rep(TRUE, n))
      rng <- range(accepted, na.rm = TRUE)
      y >= rng[1] & y <= rng[2]
    }, logical(n))
    best_inliers <- best_inliers | apply(in_range, 1, all)
  }

  best_inliers
}

# define closed-form OLS slope/intercept for y ~ t
ols_fit <- function(t, y) {
  ok <- is.finite(t) & is.finite(y)
  if (sum(ok) < 2 || length(unique(t[ok])) < 2) {
    return(c(intercept = NA_real_, slope = NA_real_))
  }
  tt <- t[ok]; yy <- y[ok]
  tm <- mean(tt); ym <- mean(yy)
  slope <- sum((tt - tm) * (yy - ym)) / sum((tt - tm)^2)
  c(intercept = ym - slope * tm, slope = slope)
}

# corrected band columns, already computed in 00_download_AquaMatch_data.Rmd
band_cols <- c("red_corr_7", "green_corr_7", "blue_corr_7",
               "nir_corr_7", "swir1_corr_7", "swir2_corr_7", "surfacetemp_corr_7")

siteSR_LS7corr_path <- file.path(aquamatch_raw_data_dir, "siteSR_DSWE1_LS7corr.parquet")

# stay lazy (arrow Dataset) through each QA stage below, logging row/site
# counts along the way without collecting into memory - only the final,
# fully-scoped query gets collect()'d, and only when the cached outputs
# below don't already exist
siteSR_full <- open_dataset(siteSR_LS7corr_path, format = "parquet")

# HUC2 basins that appear anywhere in this project's final regional output -
# a cheap, coarse pushdown pre-filter ahead of the precise HUC4/geometry
# scoping below (which needs a spatial join, not a column predicate)
regional_huc2 <- unique(siteSR_full)

row_log <- log_step(row_log, "siteSR", "0_raw_nationwide",
                    n_rows = siteSR_full %>% 
                      summarize(n = n()) %>% 
                      collect() %>% 
                      pull(n),
                    n_sites = siteSR_full %>% 
                      summarize(n = n_distinct(siteSR_id)) %>% 
                      collect() %>% 
                      pull(n))

# HUC2 basins that appear anywhere in this project's final regional output -
# a cheap, coarse pushdown pre-filter ahead of the precise HUC4/geometry
# scoping below (which needs a spatial join, not a column predicate)
regional_huc2 <- c(10, 11, 13, 14, 15, 16, 17)

siteSR_huc2 <- siteSR_full %>% filter(huc2 %in% regional_huc2)

# precise regional scoping: single-waterbody geometry QA (done above when
# `sites` was built) and the 6-state HUC4 footprint, applied here before
# collecting rather than after matching in 02a_make_matches.Rmd
siteSR_geom_huc4 <- siteSR_huc2 %>% 
  filter(siteSR_id %in% sites$siteSR_id)
row_log <- log_step(row_log, "siteSR", "1_geometry_and_HUC4_scoped",
                    n_rows = siteSR_geom_huc4 %>% 
                      summarize(n = n()) %>% 
                      collect() %>% 
                      pull(n),
                    n_sites = siteSR_geom_huc4 %>% 
                      summarize(n = n_distinct(siteSR_id)) %>% 
                      collect() %>% 
                      pull(n))

siteSR_metaqa <- siteSR_geom_huc4 %>%
  filter(pCount_dswe1 / pCount_dswe_gt0 > 0.5,
         pCount_dswe1 > 8,
         # only rows with no clouds in the buffer area
         prop_clouds == 0,
         # no rows with median temp below freezing
         flag_temp_min == 0,
         # or above 40deg
         flag_temp_max == 0,
         # filter for scenes with < 20% cloud cover
         sat_id %in% scene_metadata$sat_id)
row_log <- log_step(row_log, "siteSR", "2_metadata_scene_QA",
                    n_rows = siteSR_metaqa %>% 
                      summarize(n = n()) %>% 
                      collect() %>% 
                      pull(n),
                    n_sites = siteSR_metaqa %>% 
                      summarize(n = n_distinct(siteSR_id)) %>% 
                      collect() %>% 
                      pull(n))

if (any(!file.exists(c(siteSR_regional_filtered_path, siteSR_regional_corr_diag_path)))) {

  siteSR_regional_raw <- siteSR_metaqa %>% collect()

  siteSR_regional_dedup <- siteSR_regional_raw %>%
    # only choose one representative site per location
    filter(siteSR_id %in% representative_sites) %>%
    arrange(siteSR_id, date)

  # run per site, parallelized across site-groups; mc.cores leaves 2 cores
  # free. group_split() (not data.table) for grouping, so each group stays a
  # plain tibble the ransac function can index by column name
  set.seed(47)
  site_splits <- siteSR_regional_dedup %>%
    group_by(siteSR_id) %>%
    group_split()

  site_splits_flagged <- mclapply(
    site_splits,
    function(site_df) {
      site_df %>%
        mutate(is_band_inlier = ransac_site_multiband(date, as.matrix(pick(all_of(band_cols)))))
    },
    mc.cores = max(1, detectCores() - 2)
  )

  # collapse and clean up
  siteSR_regional_corr <- bind_rows(site_splits_flagged)
  rm(site_splits, site_splits_flagged)
  gc()

  # persist the pre-filter data (with is_band_inlier still attached) for the
  # diagnostics chunks below, since siteSR_regional_filtered_df further down
  # drops both the flagged rows and the is_band_inlier column itself
  write_parquet(siteSR_regional_corr, siteSR_regional_corr_diag_path)

  # siteSR_regional_corr (and its is_band_inlier flags) only covers
  # representative_sites - one siteSR_id per location_id. The other
  # siteSR_id(s) collapsed into each multi-site location_id (see
  # site-location-dedup above) are the same physical pixel(s), observed on
  # the same overpasses, so rather than re-running the expensive per-site
  # ransac_site_multiband() fit on them too, we reuse the representative
  # site's decision for that scene. sat_id is a scene-global id (Earth
  # Engine's system:index - see download_sceneMetadata.R), not a per-site
  # one, so the lookup must be keyed on (location_id, sat_id)
  location_lookup <- sites %>%
    select(siteSR_id, location_id)

  inlier_by_location_scene <- siteSR_regional_corr %>%
    select(siteSR_id, sat_id, is_band_inlier) %>%
    left_join(location_lookup, by = "siteSR_id") %>%
    select(location_id, sat_id, is_band_inlier) %>%
    distinct()

  # re-expand from "one representative row per location_id" back out to
  # every geometry-QA-passing siteSR_id, then attach each row's location_id's scene-level inlier
  # decision. A duplicate site's sat_id that never appeared for its
  # representative (or vice versa - e.g. one site's buffer had more
  # cloud/fewer clear pixels that scene) has no match here and comes through
  # as NA, which - like a ransac fit that never resolved - is kept below
  # rather than assumed guilty (see the filter() just below: NA passes).
  siteSR_regional_filtered_df <- siteSR_regional_raw %>%
    filter(siteSR_id %in% sites$siteSR_id) %>%
    left_join(location_lookup, by = "siteSR_id") %>%
    left_join(inlier_by_location_scene, by = c("location_id", "sat_id"))

  siteSR_regional_filtered_df %>%
    filter(is_band_inlier | is.na(is_band_inlier)) %>%
    select(-location_id, -is_band_inlier) %>%
    write_parquet(siteSR_regional_filtered_path)

  rm(siteSR_regional_raw, siteSR_regional_corr, siteSR_regional_filtered_df,
     location_lookup, inlier_by_location_scene)
}

# log the final output regardless of whether it was just written above or
# already existed from a prior run
siteSR_final <- open_dataset(siteSR_regional_filtered_path)
row_log <- log_step(row_log, "siteSR", "3_band_ransac_filter_final",
                    n_rows = siteSR_final %>% 
                      summarize(n = n()) %>% 
                      collect() %>% 
                      pull(n),
                    n_sites = siteSR_final %>% 
                      summarize(n = n_distinct(siteSR_id)) %>% 
                      collect() %>% 
                      pull(n))
```

Look at some diagnostics to make sure this makes sense. These reload the
pre-filter data from `siteSR_regional_corr_diag_path` (written above) as an
arrow Dataset, so the filtering/summarizing below stays lazy and only the
rows actually needed for a given plot get collected into memory.

```{r load-siteSR-regional-corr-diag}
siteSR_regional_corr <- open_dataset(siteSR_regional_corr_diag_path)
```

Pull sites where more than 10% of points were flagged, for a first look at
whether the filter is behaving reasonably.

```{r high-flag-sites}
high_flag_sr_sites <- siteSR_regional_corr %>%
  group_by(siteSR_id) %>%
  summarize(n = n(),
            n_flagged = sum(is_band_inlier == FALSE, na.rm = TRUE)) %>%
  collect() %>%
  mutate(frac_flagged = n_flagged/n) %>%
  filter(frac_flagged > 0.1) %>%
  arrange(-frac_flagged)
```

Spot-check a handful of sites with a large enough record, but not too dense,
such that the pattern is legible: flagged points should look like
genuine spectral excursions (residual cloud/shadow contamination, glint)
rather than a large fraction of any one site's record.

```{r spotcheck-sites}
spotcheck_sites <- siteSR_regional_corr %>%
  group_by(siteSR_id) %>%
  summarize(n = n()) %>%
  filter(between(n, 950,960)) %>%
  collect() %>%
  pull(siteSR_id)

map(.x = spotcheck_sites,
    .f = ~ print(
      siteSR_regional_corr %>%
        filter(siteSR_id == .x) %>%
        collect() %>%
        pivot_longer(cols = all_of(band_cols),
                     names_to = "band",
                     values_to = "band_val") %>%
        ggplot(.,
               aes(x = date, y = band_val, color = is_band_inlier)) +
        geom_point() +
        facet_grid(band ~ siteSR_id, scales = "free_y") +
        scale_color_manual(values = c(`TRUE` = "grey40", `FALSE` = "red")) +
        theme_bw()
    )
)
```

Some of these look like they might belong to the same waterbody or are sites
that are very close to one another, even after the de-duplication step.

Check whether flagging concentrates at reflectance extremes rather than
spreading evenly across the range of each band - the same "does this look
like real signal being cut, or generic scene noise" question asked of the
SDD-only filter above, just against the band values themselves since no
SDD/harmonized_value exists on this pre-match data. Even spreading here is
consistent with catching genuine spectral noise (cloud/shadow edges,
glint); concentration at one band's extremes would suggest the filter is
instead cutting real, legitimately bright or dark water.

Deciling requires the full column in memory (arrow's Acero engine doesn't
push `ntile()` down), so this chunk collects once into a plain tibble;
that same collected copy is reused by the decile10 chunks below it.

```{r band-tail-check}
siteSR_regional_corr_df <- siteSR_regional_corr %>% 
  collect()

band_tail_check <- band_cols %>%
  map_dfr(function(b) {
    siteSR_regional_corr_df %>%
      transmute(band = b, bin = ntile(.data[[b]], 10), is_band_inlier)
  }) %>%
  group_by(band, bin) %>%
  summarize(n = n(),
            n_flagged = sum(is_band_inlier == FALSE, na.rm = TRUE),
            n_unflagged = sum(is_band_inlier == TRUE, na.rm = TRUE),
            n_na = sum(is.na(is_band_inlier)),
            .groups = "drop") %>%
  mutate(frac_flagged = n_flagged / n,
         frac_unflagged = n_unflagged / n,
         frac_na = n_na / n) %>%
  arrange(band, bin)

print(
  band_tail_check %>%
    pivot_longer(cols = c(frac_flagged, frac_unflagged, frac_na),
                 names_to = "status", values_to = "frac") %>%
    mutate(status = recode(status,
                           frac_flagged = "flagged",
                           frac_unflagged = "not flagged",
                           frac_na = "NA")) %>%
    ggplot(aes(x = bin, y = frac, fill = status)) +
    geom_col() +
    facet_wrap(~ band, scales = "free_x") +
    labs(x = "decile of corrected band value (1 = lowest)",
         y = "fraction of observations",
         fill = NULL) +
    theme_bw()
)
```

This does skew to flagging greater outliers at higher values, but that is probably
real. We can further look into that by looking at sites where there are a lot
of data being flagged in the 90th percentile.

Spot-check sites with a large enough record (> 1000 observations) that the
pattern is legible, restricted to sites with at least one point in that
band's decile 10 (i.e. at/above the band's global 90th percentile) - a
closer look at the sites/points actually behind the decile-10 concentration
surfaced above, to see whether those points look like isolated spectral
noise or a recurring signal (e.g. seasonal bloom) the filter is cutting.

```{r decile10-spotcheck-prep}
band_p90 <- set_names(
  map_dbl(band_cols, ~ quantile(siteSR_regional_corr_df[[.x]], 0.9, na.rm = TRUE)),
  band_cols
)
band_p90_df <- tibble(band = band_cols, p90 = band_p90)

siteSR_band_long <- band_cols %>%
  map_dfr(function(b) {
    siteSR_regional_corr_df %>%
      transmute(siteSR_id, date, band = b, band_val = .data[[b]], is_band_inlier)
  })

large_sites <- siteSR_regional_corr %>%
  group_by(siteSR_id) %>%
  summarize(n = n()) %>%
  filter(n > 1000) %>%
  collect() %>%
  pull(siteSR_id)

decile10_sites_all <- siteSR_band_long %>%
  filter(siteSR_id %in% large_sites, band_val >= band_p90[band]) %>%
  distinct(siteSR_id) %>%
  pull(siteSR_id)

# this condition flags the vast majority of large sites (a decile-10 hit in
# any of 7 bands is a low bar), so plotting all of them stops being a spot
# check - sample a fixed, reproducible subset instead
set.seed(43)
decile10_spotcheck_n <- 25
decile10_sites <- if (length(decile10_sites_all) > decile10_spotcheck_n) {
  sample(decile10_sites_all, decile10_spotcheck_n)
} else {
  decile10_sites_all
}
```

One figure per site, rather than one huge multi-site `facet_grid`, so each
site's timeseries is actually readable. `r length(decile10_sites_all)` sites
meet the decile10 criterion above; we spot-check a random sample of
`r length(decile10_sites)` of them here rather than rendering all of them.

```{r decile10-spotcheck-plot}
walk(decile10_sites, function(site) {
  print(
    siteSR_band_long %>%
      filter(siteSR_id == site) %>%
      ggplot(aes(x = date, y = band_val, color = is_band_inlier)) +
      geom_point() +
      geom_hline(data = band_p90_df, aes(yintercept = p90),
                 linetype = "dashed") +
      facet_wrap(~ band, scales = "free_y", ncol = 2) +
      scale_color_manual(values = c(`TRUE` = "grey40", `FALSE` = "red")) +
      labs(title = site) +
      theme_bw()
  )
})
```

Can still see some related sites in here, but will leave as is for now.

Release the large diagnostic objects before moving on to save memory.

```{r diagnostics-cleanup}
rm(siteSR_regional_corr, siteSR_regional_corr_df, band_cols,
   high_flag_sr_sites, spotcheck_sites,
   band_tail_check, band_p90, band_p90_df,
   siteSR_band_long, large_sites,
   decile10_sites, decile10_sites_all, decile10_spotcheck_n)
gc()
```

## Row-count log

Persist the SDD and siteSR row/site counts logged at each filtering step
above so `02a_make_matches.Rmd` can append its own matching/QA steps and
render a combined retention summary.

```{r write-row-log}
write_csv(row_log, file.path(filtered_aquamatch_dir, "row_loss_log_01.csv"))
```
