---
title: "Make Matchups"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 19, 2026"
output:
  html_document:
    fig_retina: 1
---

**Pipeline step 02a · R · run from the repository root**

- **Purpose:** Match filtered SDD to siteSR within a 5-day window, drop matchups with a rain event in the gap, and keep the closest image per in situ sample.
- **Inputs:** `aquamatch_files/`: `sdd_quality_filter.parquet`, `siteSR_DSWE1_regional_band_filtered.parquet`, `huc4_filtered.gdb`, `states.gdb`, `row_loss_log_01.csv`; site metadata CSV; gridMET (cached).
- **Outputs:** `aquamatch_files/`: `filtered_regional_sdd.parquet` (modeling matchups), `regional_sdd_not_closest.parquet`, `sdd_regional_siteSR_matchups_5d.parquet`, `match_gap_precip_raw.feather`, `row_loss_log.csv`.
- **Previous:** [01 · Filter AquaMatch data](01_filter_AquaMatch_Data.Rmd) · **Next:** [02b · Pull site characteristics](02b_pull_site_characteristics.Rmd), [02c · Pull weather summaries](02c_pull_weather_summaries.Rmd), then [03 · Split data](03_split_data.ipynb)

This code creates matchups between the filtered AquaMatch SDD and siteSR
files (from `01_filter_AquaMatch_Data.Rmd`) using a 5-day window.
Both inputs are already scoped to the HUC4 basins that make up this
project's 6-state regional AOI (that scoping happens in
`01_filter_AquaMatch_Data.Rmd`, before the expensive per-site filtering
there runs), so the "Subset to the regional AOI" section below is a
labeling/sanity-check step rather than an additional filter.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

Load packages and set paths to the filtered SDD and siteSR caches from
[01](01_filter_AquaMatch_Data.Rmd) and to AquaMatch's site metadata.
`AquaMatchr` performs the matching. `climateR` and `parallel` support the
rain screen further below. `tmap` draws the maps.

```{r libraries, , message = FALSE, warning=FALSE}
# load libraries
library(AquaMatchr) # if install needed: remotes::install_github(repo = "AquaSat/AquaMatchr")
library(data.table)
library(arrow)
library(sf)
library(tmap)
library(tigris)
library(tidyverse)
library(climateR)
library(parallel)

aquamatch_raw_data_dir <- "regional_clarity/aquamatch_files/downloads"
filtered_aquamatch_dir <- "regional_clarity/aquamatch_files"

# outputs from 01_filter_AquaMatch_Data.Rmd
sdd_filtered_path <- file.path(filtered_aquamatch_dir, "sdd_quality_filter.parquet")
siteSR_regional_filtered_path <- file.path(filtered_aquamatch_dir, "siteSR_DSWE1_regional_band_filtered.parquet")
sr_sites <- file.path(aquamatch_raw_data_dir, "siteSR_collated_WQP_NWIS_sites_with_NHD_info_2025-06-04.csv")
```

## Make matches with the filtered data

Use the regionally band-filtered siteSR data (rather than the full
nationwide `unified_data`) as the input to matching, so a site flagged in
script 01 simply isn't a match candidate.

`match_siteSR_to_WQP()` only accepts `.csv`/`.feather` for its WQP input and
hard-codes `.feather` for its siteSR input - parquet isn't a supported input
format for this AquaMatchr function. Our filtered caches from script 01 are
parquet (see that script's note on preferring parquet everywhere it isn't
required to be otherwise), so we write throwaway feather copies just for
this call and delete them immediately after.

Only SDD observations that RANSAC kept (or could not evaluate) are passed
to the matcher. Every siteSR observation at the same site within ±5 days of
an SDD sample becomes a candidate matchup. The result is cached, so later
renders read it from disk.

```{r run-siteSR-wqp-match, message = FALSE, warning=FALSE}
out_file <- file.path(filtered_aquamatch_dir, "sdd_regional_siteSR_matchups_5d.parquet")

if (file.exists(out_file)) {
  sdd_match <- read_parquet(out_file)
} else {
  # throwaway feather copies, required by match_siteSR_to_WQP()'s input format
  sdd_filtered_feather <- tempfile(fileext = ".feather")
  siteSR_regional_filtered_feather <- tempfile(fileext = ".feather")
  # is_inlier == FALSE is a RANSAC-flagged SDD outlier (see
  # 01_filter_AquaMatch_Data.Rmd); NA (RANSAC never resolved, e.g. too few
  # points at that site) is kept rather than assumed guilty
  write_feather(read_parquet(sdd_filtered_path) %>%
                  filter(is_inlier | is.na(is_inlier)),
                sdd_filtered_feather)
  write_feather(read_parquet(siteSR_regional_filtered_path), siteSR_regional_filtered_feather)
  
  # Run the matchups with a 5-day window
  match_siteSR_to_WQP(
    wqp_path = sdd_filtered_feather,
    siteSR_path = siteSR_regional_filtered_feather,
    site_list_path = sr_sites,
    save_location = out_file,
    time_window = "5 days"
  )
  
  file.remove(sdd_filtered_feather, siteSR_regional_filtered_feather)
  
  sdd_match <- read_parquet(out_file)
}

```

Let's confirm we're only matching Lake data for our purposes and that the
Great Lakes were already excluded - this is a sanity check on the site
filters already applied upstream, not an additional filter.

```{r lakes-only}
lake_sdd <- sdd_match %>%
  filter(
    # select for lake/reservoir only
    grepl("Lake|Reservoir", MonitoringLocationTypeName),
    # remove the great lakes for this matching process
    !grepl("Great", MonitoringLocationTypeName)
  )

# check to see if these are the same
nrow(lake_sdd) == nrow(sdd_match)
```

## Summary Figures

Read AquaMatch's site metadata for site coordinates.

```{r site-list, message = FALSE, warning=FALSE}
sites <- read_csv(sr_sites, show_col_types = FALSE)
```

Convert the sites that have at least one matchup to spatial points and keep
those within the contiguous United States (CONUS) for mapping.

```{r conus-only, , message = FALSE, warning=FALSE}
sites_w_matches <- sites %>%
  filter(siteSR_id %in% unique(sdd_match$siteSR_id)) %>%
  st_as_sf(coords = c("WGS84_Longitude", "WGS84_Latitude"), crs = "EPSG:4326")

# let's drop the ones that are outside the CONUS for the purposes of this
conus <- suppressMessages(states(progress_bar = FALSE)) %>%
  filter(!grepl("AK|AS|MP|VI|HI|PR|GU", STUSPS))

lake_locs_CONUS <- sites_w_matches %>%
  st_transform(., st_crs(conus)) %>%
  .[conus,] %>%
  st_transform(., st_crs(sites_w_matches))
```

Map every site with at least one matchup.

```{r}
tm_shape(conus) +
  tm_borders() +
  tm_shape(lake_locs_CONUS) +
  tm_dots()
```

Count matchups per site and attach the counts to the site locations.

```{r summary-by-site}
# summarize by number of matches per site
five_day_summary <- sdd_match %>%
  summarize(n = n(),
            .by = siteSR_id)

# add in spatial data
lake_locs_with_matches <- lake_locs_CONUS %>%
  filter(siteSR_id %in% unique(sdd_match$siteSR_id)) %>%
  left_join(., five_day_summary)
```

Plot site locations colored by number of matchups (all Landsat missions
combined).

```{r plot-all-missions}
# plot all missions
tm_shape(conus) +
  tm_polygons() +
  tm_shape(lake_locs_with_matches, bbox = conus) +
  tm_dots("n",
          fill.scale = tm_scale(breaks = c(1, 10, 25, 50, 75, 100,
                                           500, 1000, 1500),
                                values = "viridis"),
          size = 0.5,
          fill_alpha = 0.5) +
  tm_layout(legend.position = tm_pos_in("right", "bottom"),
            legend.bg.color = "white",
            legend.frame = TRUE)
```

Repeat the map for each Landsat mission separately. Each panel title gives
that mission's total matchups and sites, which shows how the record is
distributed across sensors.

```{r summary-by-mission}
# and per site per mission
five_day_summary_sensor <- sdd_match %>%
  summarize(n = n(),
            .by = c(siteSR_id, mission))

# add spatial data and label info
lake_locs_by_mission <- lake_locs_CONUS %>%
  select(siteSR_id, geometry) %>%
  right_join(five_day_summary_sensor, by = "siteSR_id") %>%
  mutate(mission = factor(mission,
                          levels = c("LT04", "LT05", "LE07", "LC08", "LC09"),
                          labels = c("Landsat 4", "Landsat 5", "Landsat 7", "Landsat 8", "Landsat 9")))%>%
  arrange(n)

# Compute totals per mission before faceting
mission_stats <- lake_locs_by_mission %>%
  st_drop_geometry() %>%
  summarize(total_matches = sum(n),
            total_sites = n_distinct(siteSR_id),
            .by = mission)

# and prep for viz
lake_locs_by_mission <- lake_locs_by_mission %>%
  left_join(mission_stats, by = "mission") %>%
  mutate(mission_label = paste0(mission, ", ",
                                format(total_matches, big.mark = ","), " matches, ",
                                format(total_sites, big.mark = ","), " sites")) %>%
  arrange(n)

# plot!
tm_shape(conus) +
  tm_polygons() +
  tm_shape(lake_locs_by_mission, bbox = conus) +
  tm_dots("n",
          fill.scale = tm_scale(values = "viridis"),
          fill.legend = tm_legend(position = tm_pos_on_top(pos.h = "right", pos.v = "bottom"),
                                  bg.color = "white",
                                  frame = TRUE, text.size = 1),
          size = 0.5,
          fill_alpha = 0.5) +
  tm_facets_wrap(by = "mission_label", ncol = 2, nrow = 3)
```

## Subset to the regional AOI

`siteSR` was already scoped to our 6-state region's HUC4 footprint upstream
in `01_filter_AquaMatch_Data.Rmd` (see that script's "Regional AOI" section)
- there's no regional-context-dependent processing between that scoping and
here, so every match in `sdd_match` should already fall inside it. We read
back the HUC4/state boundaries 01 already computed and wrote out, rather
than re-running the HUC4 download/overlap calculation, and use them here
only to label each site with its HUC4 (for the summaries below) and as a
sanity check on the upstream scoping.

```{r read-cached-boundaries, message = FALSE, warning=FALSE}
# standardize the geometry column name to "geometry" - the OpenFileGDB
# driver renames it to "SHAPE" on write, regardless of the name it had
# in 01_filter_AquaMatch_Data.Rmd before export
standardize_geom_col <- function(x) {
  geom_col <- attr(x, "sf_column")
  if (geom_col != "geometry") {
    names(x)[names(x) == geom_col] <- "geometry"
    st_geometry(x) <- "geometry"
  }
  x
}

HUC4_filtered <- st_read(file.path(filtered_aquamatch_dir, "huc4_filtered.gdb"),
                         layer = "huc4_filtered", quiet = TRUE) %>%
  standardize_geom_col()
states <- st_read(file.path(filtered_aquamatch_dir, "states.gdb"),
                  layer = "states", quiet = TRUE) %>%
  standardize_geom_col()
```

Label each matched site with the HUC4 it falls in and carry that label onto
the matchups. The printed message confirms that every matched site falls
inside the regional footprint.

```{r huc4-label-join}
# add huc info - this is a labeling join (and a sanity check on the
# upstream regional scoping), not an additional filter: `left = FALSE` drops
# any matched site that somehow falls outside every regional HUC4, which
# shouldn't happen given the upstream scoping
regional_sites <- sites_w_matches %>%
  st_transform(st_crs(HUC4_filtered)) %>%
  st_join(HUC4_filtered %>% select(huc4), left = FALSE)

cat(n_distinct(regional_sites$siteSR_id), "of",
    n_distinct(sites_w_matches$siteSR_id),
    "matched sites fall within the regional HUC4 footprint",
    "(should be all of them, since siteSR was already scoped to this",
    "footprint upstream in 01_filter_AquaMatch_Data.Rmd)\n")

# add sdd info back, carrying along each site's HUC4 label
regional_sdd <- sdd_match %>%
  filter(siteSR_id %in% regional_sites$siteSR_id) %>%
  left_join(
    regional_sites %>%
      st_drop_geometry() %>%
      select(siteSR_id, HUC4 = huc4) %>%
      # guard against a site sitting exactly on a shared HUC4 boundary,
      # which could otherwise match more than one polygon in the spatial join
      distinct(siteSR_id, .keep_all = TRUE),
    by = "siteSR_id"
  )
```

## Screen for heavy rain events between image and field dates

For each candidate match (every SDD/siteSR pair within the 5-day window,
before any closest-in-time selection) we check to see if there was a heavy single-day rain
event between the image date and the field (SDD) date (in either order). If
so, the reflectance and the target likely describe genuinely different water
states, and the matchup shouldn't be trusted regardless of which one came
first. NOAA/EPA's "heavy precipitation day" cut of 1 inch (25.4 mm) in 24
hours, peak single-day intensity within the gap (not cumulative gap-total,
which would conflate a genuine storm with a longer gap simply having more
days to accumulate ordinary rain).

This screens **candidates**, before the closest-in-time selection below, not
after: filtering the already-deduped closest-match dataset would drop an
observation outright whenever its single nearest candidate image happened to
be rain-contaminated, even when a perfectly good, only slightly-further-away,
uncontaminated image was also available in the window. Screening first lets
closest-in-time selection fall back to that next-best candidate instead.

Needs raw daily gridMET precipitation for every candidate site, over each
site's full candidate-date range - a broader, coarser need than
`02c_pull_weather_summaries.Rmd`'s antecedent-window feature fetch (which only
needs narrow 1-30 day windows ending at each *already-matched* row's date),
so this is its own fetch rather than a reuse of that cache. Also breaks a
real circularity: `02c_pull_weather_summaries.Rmd` fetches weather for the
*output* of this script, so it isn't available yet at this point in the
pipeline.

```{r rain-check-weather-fetch, message=FALSE, warning=FALSE}
HEAVY_RAIN_MM <- 25.4  # NOAA/EPA "heavy precipitation day" threshold (1 inch/24hr)

regional_sdd <- regional_sdd %>%
  mutate(date = as.Date(date),
         field_date = as.Date(harmonized_local_time),
         gap_start = pmin(date, field_date),
         gap_end = pmax(date, field_date))

rain_check_weather_raw_path <- file.path(filtered_aquamatch_dir, "match_gap_precip_raw.feather")

if (file.exists(rain_check_weather_raw_path)) {
  gap_precip_raw <- read_feather(rain_check_weather_raw_path)
} else {
  site_ranges <- regional_sdd %>%
    left_join(sites %>% select(siteSR_id, WGS84_Latitude, WGS84_Longitude), by = "siteSR_id") %>%
    summarize(min_date = min(gap_start) - 1, max_date = max(gap_end),
              lat = first(WGS84_Latitude), lon = first(WGS84_Longitude), .by = siteSR_id)
  cat(sprintf("[%s] fetching gridMET precip for %s sites\n", format(Sys.time(), "%H:%M:%S"), nrow(site_ranges)))

  fetch_one <- function(i, site_ranges) {
    row <- site_ranges[i, ]
    fetch <- function() {
      pt <- sf::st_as_sf(data.frame(lon = row$lon, lat = row$lat), coords = c("lon", "lat"), crs = 4326)
      res <- getGridMET(AOI = pt, varname = "pr",
                        startDate = as.character(row$min_date), endDate = as.character(row$max_date))
      res$siteSR_id <- row$siteSR_id
      res
    }
    tryCatch(fetch(), error = function(e) {
      Sys.sleep(2)
      tryCatch(fetch(), error = function(e2) data.frame(siteSR_id = row$siteSR_id, error = conditionMessage(e2)))
    })
  }

  t0 <- Sys.time()
  cl <- makeCluster(8, type = "PSOCK")
  clusterEvalQ(cl, { suppressMessages(library(climateR)); suppressMessages(library(sf)) })
  clusterExport(cl, c("site_ranges", "fetch_one"))
  results <- parLapply(cl, seq_len(nrow(site_ranges)), fetch_one, site_ranges = site_ranges)
  stopCluster(cl)
  cat(sprintf("[%s] fetch took %.1f min\n", format(Sys.time(), "%H:%M:%S"),
              as.numeric(difftime(Sys.time(), t0, units = "mins"))))

  ok <- Filter(function(x) !"error" %in% names(x), results)
  failed <- Filter(function(x) "error" %in% names(x), results)
  cat(sprintf("[%s] succeeded: %s sites, failed: %s sites\n", format(Sys.time(), "%H:%M:%S"), length(ok), length(failed)))
  if (length(failed) > 0) {
    write_csv(bind_rows(failed), file.path(filtered_aquamatch_dir, "match_gap_precip_fetch_failures.csv"))
  }

  gap_precip_raw <- bind_rows(ok) %>% mutate(date = as.Date(date))
  write_feather(gap_precip_raw, rain_check_weather_raw_path)
}
```

For each candidate matchup, find the largest single-day precipitation total
between the earlier and later of the image and field dates (excluding the
earlier date itself). Candidates with a day at or above 25.4 mm are removed.
A `data.table` interval join performs this lookup in one pass over all
candidates.

```{r rain-check-apply}
# peak single-day precip strictly BETWEEN gap_start and gap_end (exclusive of
# gap_start itself, the "day of" whichever event is earlier) - a data.table
# non-equi (interval) join, not a row-by-row filter (the latter would be an
# O(n_candidates x n_weather_rows) scan)
regional_sdd <- regional_sdd %>% mutate(cand_id = row_number())
candidates_dt <- as.data.table(regional_sdd %>%
                                 select(cand_id, siteSR_id, gap_start, gap_end))
wx_dt <- as.data.table(gap_precip_raw %>% select(siteSR_id, date, pr))
setkey(wx_dt, siteSR_id, date)

joined <- wx_dt[candidates_dt, on = .(siteSR_id, date > gap_start, date <= gap_end),
                nomatch = NA, allow.cartesian = TRUE]
gap_max <- joined[, .(gap_max_precip_mm = max(pr, na.rm = TRUE)), by = cand_id]
gap_max[!is.finite(gap_max_precip_mm), gap_max_precip_mm := 0]

regional_sdd <- regional_sdd %>%
  left_join(as.data.frame(gap_max), by = "cand_id") %>%
  mutate(gap_max_precip_mm = coalesce(gap_max_precip_mm, 0))

n_flagged <- sum(regional_sdd$gap_max_precip_mm >= HEAVY_RAIN_MM)
cat(sprintf("[%s] %s / %s candidate matches flagged for a heavy single-day rain event (>= %.1f mm) between image and field dates - excluded before closest-match selection\n",
            format(Sys.time(), "%H:%M:%S"), n_flagged, nrow(regional_sdd), HEAVY_RAIN_MM))

regional_sdd <- regional_sdd %>%
  filter(gap_max_precip_mm < HEAVY_RAIN_MM) %>%
  select(-field_date, -gap_start, -gap_end, -cand_id, -gap_max_precip_mm)
```

## QA the data

Make sure we're using the observation closest in time to the measurement.
This assures we're not using multiple satellites for the same data point in our
training dataset. We will also export those that are dismissed at this point as a 
possible additional test set.

```{r closest-obs}
# grab the observation closes in date to a satellite flyover, from the
# rain-screened candidate pool built above
reg_sdd_closest <- regional_sdd %>%
  arrange(abs(time_diff), mission) %>%
  slice(1, .by = c("subgroup_id"))

# let's also do an anti-join here of the other observations so that we have an
# additional validation dataset to play with.
reg_sdd_additional <- anti_join(regional_sdd, reg_sdd_closest)

write_parquet(reg_sdd_additional,
              file.path(filtered_aquamatch_dir, "regional_sdd_not_closest.parquet"))
```

## Visualize data

Let's look at some longer records to do a sanity check to make sure that all
the filters worked correctly.

```{r site-summary}
# summarize by site for plotting
regional_site_summary <- reg_sdd_closest %>%
  st_as_sf(coords = c("lon", "lat"),
           crs = "EPSG:4326") %>%
  summarize(n = n(), .by = c(siteSR_id, geometry)) %>%
  arrange(n)

# look at some of the records with a lot of data
regional_long_record <- regional_site_summary %>%
  filter(n > 100) %>%
  left_join(., reg_sdd_closest)

ggplot(regional_long_record, aes(x = date, y = harmonized_value)) +
  geom_point() +
  facet_wrap(siteSR_id ~ .) +
  theme_bw()
```

Map the number of final matchups at each site, overlaid on the regional
HUC4 boundaries.

```{r plot-sites-spatial}
tm_shape(HUC4_filtered) +
  tm_shape(states %>%
             st_transform(st_crs(HUC4_filtered))) +
  tm_polygons() +
  tm_shape(HUC4_filtered) +
  tm_polygons(fill_alpha = 0) +
  tm_shape(regional_site_summary %>%
             st_transform(st_crs(HUC4_filtered))) +
  tm_dots("n",
          fill.scale = tm_scale_continuous(values = "viridis"))
```

Summarize the final matchups by HUC4. Basins with no matchups are outlined
but left unfilled.

```{r plot-huc-summary}
regional_huc_summary <- reg_sdd_closest %>%
  st_as_sf(coords = c("lon", "lat"),
           crs = "EPSG:4326") %>%
  st_transform(., st_crs(HUC4_filtered)) %>%
  st_join(HUC4_filtered, .) %>%
  summarize(n = sum(!is.na(siteSR_id)), .by = c(huc4, geometry))


tm_shape(HUC4_filtered) +
  tm_shape(states %>%
             st_transform(st_crs(HUC4_filtered))) +
  tm_polygons() +
  tm_shape(HUC4_filtered) +
  tm_polygons(fill_alpha = 0) +
  # drop 0-count HUC4s so they're left uncolored (outline only) instead of
  # taking the low end of the viridis scale
  tm_shape(regional_huc_summary %>% filter(n > 0)) +
  tm_polygons("n",
              fill.scale = tm_scale_continuous(values = "viridis"),
              fill_alpha = 0.5)

```

## Data Retention Summary

How much data survived each filtering step, from the raw nationwide SDD and
siteSR inputs down to the final matched, closest-observation training set?
`01_filter_AquaMatch_Data.Rmd` logged row/site counts through its SDD and
siteSR filtering steps (`row_loss_log_01.csv`); we append this script's
matching and closest-observation QA steps here. Every step is already
scoped to this project's 6-state HUC4 region except each pipeline's first
("raw") row.

```{r row-log-combine}
row_log <- read_csv(file.path(filtered_aquamatch_dir, "row_loss_log_01.csv"),
                    show_col_types = FALSE) %>%
  bind_rows(
    tibble(pipeline = "matched", step = "0_5day_window_match",
           n_rows = nrow(sdd_match), n_sites = n_distinct(sdd_match$siteSR_id)),
    # regional_sdd was reassigned to the rain-screened candidate pool above
    tibble(pipeline = "matched", step = "1_heavy_rain_screen",
           n_rows = nrow(regional_sdd), n_sites = n_distinct(regional_sdd$siteSR_id)),
    tibble(pipeline = "matched", step = "2_closest_observation_final",
           n_rows = nrow(reg_sdd_closest), n_sites = n_distinct(reg_sdd_closest$siteSR_id))
  )

write_csv(row_log, file.path(filtered_aquamatch_dir, "row_loss_log.csv"))
```

Tabulate the log: for each pipeline (SDD, siteSR, and matched), the rows
and sites remaining after each step, as a percentage of the previous step
and of the pipeline's starting count.

```{r row-log-table}
row_log_display <- row_log %>%
  mutate(pipeline = factor(pipeline, levels = c("SDD", "siteSR", "matched"))) %>%
  arrange(pipeline, step) %>%
  group_by(pipeline) %>%
  mutate(step_label = step %>%
           str_remove("^[0-9]+[a-z]?_") %>%
           str_replace_all("_", " ") %>%
           str_to_sentence(),
         pct_of_raw = round(100 * n_rows / first(n_rows), 1),
         pct_of_previous = round(100 * n_rows / lag(n_rows), 1)) %>%
  ungroup() %>%
  select(pipeline, step_label, n_rows, n_sites, pct_of_previous, pct_of_raw)

knitr::kable(row_log_display,
             col.names = c("Pipeline", "Step", "Total Rows", "Total Sites",
                           "% of previous step retained", "% of pipeline's first step retained"),
             format.args = list(big.mark = ",")) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  # merge consecutive same-pipeline cells into one spanning cell, so the
  # table reads as three grouped blocks (SDD/siteSR/matched) rather than a
  # flat list repeating the pipeline name on every row
  kableExtra::collapse_rows(columns = 1, valign = "top")
```

The same retention counts as a bar chart, one panel per pipeline, with steps
ordered top to bottom.

```{r row-log-figure, fig.height=8}
row_log_display %>%
  # fct_inorder() alone would put the first (largest) step at the bottom
  # after coord_flip() - reverse it so bars read largest-to-smallest top-down
  mutate(step_label = fct_rev(fct_inorder(step_label))) %>%
  ggplot(aes(x = step_label, y = n_rows)) +
  geom_col() +
  geom_text(aes(label = scales::comma(n_rows)), hjust = -0.05, size = 3) +
  facet_wrap(~ pipeline, scales = "free", ncol = 1) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
  labs(x = NULL, y = "rows remaining",
       title = "Data retention through the filtering pipeline") +
  ggthemes::theme_few() +
  theme(legend.position = "none")
```

## Export data

Write the final modeling matchups: one closest-in-time, rain-screened
image per SDD observation. This file is the input to the site
characteristics and weather pulls and to
[03 · Split data](03_split_data.ipynb).

```{r export}
write_parquet(reg_sdd_closest,
              file.path(filtered_aquamatch_dir, 
                        "filtered_regional_sdd.parquet"))
```
