---
title: "Pull Site Characteristics"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 16, 2026"
output: html_document
# Quarto site build: show code only, don't re-run the fetches
execute:
  eval: false
---

**Pipeline step 02b · R · run from `regional_clarity/` · output is cached; re-run only when the site set changes; not executed when the site is built**

- **Purpose:** Fetch static per-site features: point elevation and LakeCat catchment land-cover metrics.
- **Inputs:** `aquamatch_files/filtered_regional_sdd.parquet`; site metadata CSV; elevation and LakeCat APIs.
- **Outputs:** `aquamatch_files/site_characteristics.feather`.
- **Previous:** [02a · Make matchups](02a_make_matches.Rmd) · **Next:** [02c · Pull weather summaries](02c_pull_weather_summaries.Rmd)

This script pulls static site/catchment characteristics — point elevation and
LakeCat catchment land-cover metrics — for a set of AquaMatch sites, and
writes a small per-site lookup table for use as model features (joined in
`03_split_data.ipynb`).

The lookup table only depends on which sites are in the matchup set, so
it is cached. Run this step when building the pipeline from scratch, and
again whenever changes in `01`/`02a` alter the site set. Otherwise the
cached table can be reused.

Variable choice follows Topp et al. 2021 (*Environ. Res. Lett.* 16 055025),
which paired optical variables with LakeCat catchment metrics
(imperviousness, urban/forest/cropland/wetland land cover, all from the 2006
land-cover snapshot) for a national Secchi-depth model.

## Requirements for the input site table

Any AquaMatch-derived (or AquaMatch-shaped) table with one row per
observation and, at minimum, the columns below will work as
`sites_input_path` — this script `distinct()`s down to one row per site
before fetching anything:

- `siteSR_id` — site identifier
- `lat`, `lon` — WGS84 coordinates

`site_metadata_path` is AquaMatch's own raw site list (the
`siteSR_collated_WQP_NWIS_sites_with_NHD_info_*.csv` produced upstream of
this project, e.g. by `AquaMatch_siteSR_WQP`'s `add_NHD_waterbody_to_sites.R`
step) — it carries the NHDPlusV2 waterbody linkage (`wb_nhd_id`) each site
was already matched to, which is what LakeCat is keyed on.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

Load the packages for spatial handling (`sf`), elevation lookups (`elevatr`),
and LakeCat queries (`StreamCatTools`), define a small timestamped logging
helper, and set the input and output paths.

```{r libraries, message=FALSE, warning=FALSE}
library(arrow)
library(dplyr)
library(readr)
library(sf)
library(elevatr)
library(StreamCatTools)

log <- function(...) cat(sprintf("[%s] ", format(Sys.time(), "%H:%M:%S")), sprintf(...), "\n")

# -- inputs (AquaMatch-shaped; see "Requirements" above) --
# this branch's canonical filtered/matched site set (03_split_data.ipynb's
# own input, from 02a_make_matches.Rmd)
sites_input_path <- "aquamatch_files/filtered_regional_sdd.parquet"
# AquaMatch's raw site metadata, carrying the NHDPlusV2 waterbody linkage
site_metadata_path <- "aquamatch_files/downloads/siteSR_collated_WQP_NWIS_sites_with_NHD_info_2025-06-04.csv"

# -- output --
output_path <- "aquamatch_files/site_characteristics.feather"
```

## Unique sites in the input set

Reduce the observation-level input to one row per site, then join AquaMatch's
site metadata to attach each site's NHD waterbody identifier (`wb_nhd_id`),
source, name, and area. The log reports how many sites have a usable
waterbody identifier for the LakeCat join.

```{r unique-sites}
final <- read_parquet(sites_input_path, col_select = c("siteSR_id", "lat", "lon"))
sites_unique <- distinct(final, siteSR_id, lat, lon)
log("n unique sites: %s", nrow(sites_unique))

sites_meta <- read_csv(site_metadata_path, show_col_types = FALSE) %>%
  select(siteSR_id, wb_nhd_id, wb_nhd_source, wb_gnis_name, wb_areasqkm)
sites_unique <- sites_unique %>% left_join(sites_meta, by = "siteSR_id")
log("sites with a valid wb_nhd_id (comid): %s / %s",
    sum(!is.na(sites_unique$wb_nhd_id)), nrow(sites_unique))
```

`wb_nhd_id` is treated below as an NHDPlusV2 comid — safe to join straight
into LakeCat with no crosswalk, *provided* every site's `wb_nhd_source` is
`"NHDPlusv2"` (the alternative, `Permanent_Identifier`, only fires for
non-CONUS HUC4s >= 1900). This was true for all 1,525 sites in the original
research pull; re-checked here every run rather than assumed, since the
active site set can change as `01`/`02a` are reworked.

```{r verify-nhd-source}
stopifnot("some sites resolve to a non-NHDPlusv2 waterbody source - LakeCat join below is not safe as-is" =
            all(sites_unique$wb_nhd_source == "NHDPlusv2" | is.na(sites_unique$wb_nhd_source)))
```

## 1. Elevation (point, AWS terrain tiles)

Look up the elevation at each site's coordinates from the AWS terrain tiles
at zoom level 9, which gives roughly 300 m ground resolution at mid-latitudes. Lake
surface elevation serves as a proxy for climate and landscape position (e.g., high-elevation
oligotrophic lakes versus low-elevation reservoirs).

```{r fetch-elevation}
log("fetching elevation for all sites")
pts_sf <- st_as_sf(sites_unique, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
t0 <- Sys.time()
elev <- get_elev_point(pts_sf, src = "aws", z = 9)
log("elevation fetch took %.1f min", as.numeric(difftime(Sys.time(), t0, units = "mins")))

elev_df <- st_drop_geometry(elev) %>% select(siteSR_id, elevation_m = elevation)
sites_unique <- sites_unique %>% left_join(elev_df, by = "siteSR_id")
log("elevation summary:"); print(summary(sites_unique$elevation_m))
```

## 2. LakeCat catchment characteristics

Land-cover-focused set, matching Topp et al.'s variable categories:
imperviousness, urban/forest/cropland/wetland land cover (2006, static, same
year Topp used), plus catchment area (LakeCat provides this for free
alongside any metric query).

Comids are requested in batches of 200 to keep each request URL a
manageable length. The individual land-cover classes are then summed into
combined forest, wetland, and urban percentages, and the results are joined
back to the sites by comid.

```{r fetch-lakecat}
log("fetching LakeCat catchment metrics")
lakecat_metrics <- paste(c(
  "pctimp2006", "pcturbhi2006", "pcturblo2006", "pcturbmd2006", "pcturbop2006",
  "pctconif2006", "pctdecid2006", "pctmxfst2006",
  "pcthbwet2006", "pctwdwet2006", "pctcrop2006"
), collapse = ",")

comids <- unique(na.omit(sites_unique$wb_nhd_id))
log("n unique comids: %s", length(comids))

# batch in chunks to keep the request URL a reasonable length
chunk_size <- 200
chunks <- split(comids, ceiling(seq_along(comids) / chunk_size))
lakecat_raw <- purrr::map_dfr(chunks, function(ch) {
  lc_get_data(comid = paste(ch, collapse = ","), metric = lakecat_metrics,
              aoi = "catchment", showAreaSqKm = TRUE)
})
log("LakeCat rows returned: %s / %s comids requested", nrow(lakecat_raw), length(comids))

lakecat <- lakecat_raw %>%
  mutate(comid = as.character(comid),
         pct_forest_2006 = pctconif2006cat + pctdecid2006cat + pctmxfst2006cat,
         pct_wetland_2006 = pcthbwet2006cat + pctwdwet2006cat,
         pct_urban_2006 = pcturbhi2006cat + pcturblo2006cat + pcturbmd2006cat + pcturbop2006cat) %>%
  select(comid, catchment_area_sqkm = catareasqkm,
         pct_impervious_2006 = pctimp2006cat, pct_urban_2006, pct_forest_2006,
         pct_cropland_2006 = pctcrop2006cat, pct_wetland_2006)

sites_unique <- sites_unique %>% left_join(lakecat, by = c("wb_nhd_id" = "comid"))
log("sites with LakeCat data: %s / %s", sum(!is.na(sites_unique$pct_impervious_2006)), nrow(sites_unique))
```

## Export

Write the per-site lookup table. `wb_nhd_source` is dropped because it was
only needed for the check above. [03 · Split data](03_split_data.ipynb)
joins this table to the matchups by `siteSR_id`.

```{r export}
write_feather(sites_unique %>% select(-wb_nhd_source), output_path)
log("wrote %s (%s cols, %s rows)", output_path, ncol(sites_unique) - 1, nrow(sites_unique))
```
