---
title: "Pull Weather Summaries"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 16, 2026"
output: html_document
# Quarto site build: show code only, don't re-run the fetches
execute:
  eval: false
---

**Pipeline step 02c · R · run from `regional_clarity/` · output is cached; re-run only when the site set or date range changes; not executed when the site is built**

- **Purpose:** Fetch daily gridMET weather per site and summarize it into 1/3/7/30-day antecedent-weather features.
- **Inputs:** `aquamatch_files/filtered_regional_sdd.parquet`; gridMET.
- **Outputs:** `aquamatch_files/`: `weather_raw.feather`, `weather_summaries.feather`.
- **Previous:** [02b · Pull site characteristics](02b_pull_site_characteristics.Rmd) · **Next:** [03 · Split data](03_split_data.ipynb)

This script pulls daily gridMET weather (precipitation, max/min/mean
temperature, solar radiation) for a set of AquaMatch sites over each site's
own observed date range, then summarizes it into "previous N days" rolling
antecedent-weather features for use as model features (joined in
`03_split_data.ipynb`).

The summaries only depend on which sites are in the matchup set and the dates
they cover, so they are cached. Run this step when building the pipeline
from scratch, and again whenever changes in `01`/`02a` alter the site set or
date range. Otherwise the cached tables can be reused. This step reads
02a's output directly, so it does not need to run after
[02b](02b_pull_site_characteristics.Rmd); the two are numbered in sequence
only for convenience.

Weather comes from gridMET via `climateR::getGridMET()` — public, no auth
token needed (the alternative, NLDAS via Earthdata Login, needs a token this
project doesn't have configured). Each site is fetched individually over
just its own observed date range (plus a lag-window buffer), not the full
gridMET record, since most sites don't need more than that.

## Requirements for the input site table

Any AquaMatch-derived (or AquaMatch-shaped) table with one row per
observation and, at minimum, the columns below will work as
`sites_input_path`:

- `siteSR_id` — site identifier
- `lat`, `lon` — WGS84 coordinates
- `date` — satellite image date
- `harmonized_local_time` — field sample date/time (used only for its date)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

Load packages, define a timestamped logging helper, and set the run
parameters: the input and output paths, the antecedent windows to summarize
(1, 3, 7, and 30 days), the gridMET variables to fetch (precipitation, daily
maximum and minimum temperature, and shortwave radiation), and the number of
parallel workers.

```{r libraries, message=FALSE, warning=FALSE}
library(arrow)
library(dplyr)
library(sf)
library(readr)
library(parallel)
library(climateR)
library(data.table)

log <- function(...) cat(sprintf("[%s] ", format(Sys.time(), "%H:%M:%S")), sprintf(...), "\n")

# -- input (AquaMatch-shaped; see "Requirements" above) --
# this branch's canonical filtered/matched site set (03_split_data.ipynb's
# own input, from 02a_make_matches.Rmd)
sites_input_path <- "aquamatch_files/filtered_regional_sdd.parquet"

# -- outputs --
output_raw_path <- "aquamatch_files/weather_raw.feather"
output_summary_path <- "aquamatch_files/weather_summaries.feather"
output_failures_path <- "aquamatch_files/weather_fetch_failures.csv"

MAX_WINDOW_DAYS <- 30  # longest antecedent window summarized below
WINDOWS <- c(1, 3, 7, 30)
VARS <- c("pr", "tmmx", "tmmn", "srad")
N_WORKERS <- 8
```

## Per-site date ranges

The match window between satellite image and field SDD sample is +-5 days,
and which side leads is close to a 50/50 split across the dataset. "Previous
N days before the image date" only covers the image-to-field gap when the
field sample came *first* — when the image comes first, that window never
reaches forward to the field date at all, silently missing real antecedent
conditions for roughly half the dataset. Anchoring each site's fetch range on
`max(image_date, field_date)` (widened by `MAX_WINDOW_DAYS + 5` days on the
early end) avoids that, and matches `03_split_data.ipynb`'s own
`anchor_date` logic for joining these summaries back onto the modeling
dataset.

```{r site-ranges}
sites_raw <- read_parquet(sites_input_path,
                          col_select = c("siteSR_id", "lat", "lon", "date", "harmonized_local_time"))
sites_raw <- sites_raw %>%
  mutate(date = as.Date(date),
         field_date = as.Date(harmonized_local_time),
         anchor_date = pmax(date, field_date))

site_ranges <- sites_raw %>%
  group_by(siteSR_id, lat, lon) %>%
  summarize(min_date = min(date) - (MAX_WINDOW_DAYS + 5), max_date = max(anchor_date), .groups = "drop")
log("fetching gridMET for %s sites", nrow(site_ranges))
```

## Fetch gridMET

`mclapply` (fork-based parallelism) crashes on macOS here — `climateR`'s
curl/geospatial dependencies touch Objective-C runtime state that isn't
fork-safe ("may have been in progress in another thread when fork() was
called ... Crashing instead"). PSOCK workers (genuinely separate processes,
not forks) avoid that.

`fetch_one()` requests the gridMET time series for a single site over its
date range and tags the result with the site ID. A failed request is retried
once after a two-second pause; a site that fails twice returns a one-row
record of the error message rather than stopping the whole run.

```{r fetch-function}
fetch_one <- function(i, site_ranges) {
  row <- site_ranges[i, ]
  fetch <- function() {
    pt <- sf::st_as_sf(data.frame(lon = row$lon, lat = row$lat),
                        coords = c("lon", "lat"), crs = 4326)
    res <- getGridMET(AOI = pt, varname = VARS,
                       startDate = as.character(row$min_date),
                       endDate = as.character(row$max_date))
    res$siteSR_id <- row$siteSR_id
    res
  }
  tryCatch(fetch(), error = function(e) {
    Sys.sleep(2)  # one retry after a short pause - transient network/server hiccups
    tryCatch(fetch(), error = function(e2) {
      data.frame(siteSR_id = row$siteSR_id, error = conditionMessage(e2))
    })
  })
}
```

Run `fetch_one()` for every site across a PSOCK cluster. Successful fetches
are stacked and saved as the raw daily weather table. Any sites that failed
both attempts are written to `weather_fetch_failures.csv` so they can be
re-requested later.

```{r fetch-run}
t0 <- Sys.time()
cl <- makeCluster(N_WORKERS, type = "PSOCK")
clusterEvalQ(cl, { suppressMessages(library(climateR)); suppressMessages(library(sf)) })
clusterExport(cl, c("site_ranges", "VARS", "fetch_one"))
results <- parLapply(cl, seq_len(nrow(site_ranges)), fetch_one, site_ranges = site_ranges)
stopCluster(cl)
log("fetch took %.1f min", as.numeric(difftime(Sys.time(), t0, units = "mins")))

ok <- Filter(function(x) !"error" %in% names(x), results)
failed <- Filter(function(x) "error" %in% names(x), results)
log("succeeded: %s sites, failed: %s sites", length(ok), length(failed))

if (length(failed) > 0) {
  write_csv(bind_rows(failed), output_failures_path)
  log("failed site IDs written to %s", output_failures_path)
}

weather_raw <- bind_rows(ok)
log("total weather rows: %s", nrow(weather_raw))
write_feather(weather_raw, output_raw_path)
```

## Summarize into rolling antecedent-weather features

Each day is summarized over rolling windows that end the day *before* that
date: total precipitation, maximum of daily highs, mean of daily means,
minimum of daily lows, and mean solar radiation. gridMET units are first
converted (temperature from Kelvin to °C). Each site's series is padded to
one row per calendar day so the fixed-width `data.table::froll*` windows are
always measured in days. The dates are then shifted forward by one so that
the value on a given date describes only the days before it.

```{r summarize}
weather <- read_feather(output_raw_path)
log("loaded %s weather rows across %s sites", nrow(weather), n_distinct(weather$siteSR_id))

# units: gridMET pr is mm, tmmx/tmmn are Kelvin, srad is W/m^2
weather <- weather %>%
  mutate(date = as.Date(date),
         precip_mm = pr,
         tmax_degC = tmmx - 273.15,
         tmin_degC = tmmn - 273.15,
         tmean_degC = (tmax_degC + tmin_degC) / 2,
         srad_Wm2 = srad) %>%
  select(siteSR_id, date, precip_mm, tmax_degC, tmin_degC, tmean_degC, srad_Wm2)

setDT(weather)
setorder(weather, siteSR_id, date)

site_splits <- split(weather, by = "siteSR_id", keep.by = TRUE)

make_summary_one_site <- function(df) {
  site_id <- df$siteSR_id[1]
  # frollsum/frollmean/frollapply assume a regular series - gridMET
  # shouldn't have gaps, but guard anyway with one row per calendar day
  full_days <- data.table(date = seq(min(df$date), max(df$date), by = "1 day"))
  df <- merge(full_days, df[, -"siteSR_id"], by = "date", all.x = TRUE)
  df[, siteSR_id := site_id]

  for (w in WINDOWS) {
    df[[paste0("precip_mm_prev", w)]] <- frollsum(df$precip_mm, n = w)
    df[[paste0("tmax_degC_prev", w)]] <- frollapply(df$tmax_degC, N = w, FUN = max)
    df[[paste0("tmean_degC_prev", w)]] <- frollmean(df$tmean_degC, n = w)
    df[[paste0("tmin_degC_prev", w)]] <- frollapply(df$tmin_degC, N = w, FUN = min)
    df[[paste0("srad_Wm2_prev", w)]] <- frollmean(df$srad_Wm2, n = w)
  }
  # shift so window w on `date` covers the w days strictly BEFORE date
  df[, date := date + 1]
  df
}

log("computing rolling summaries across %s sites", length(site_splits))
t0 <- Sys.time()
summaries <- rbindlist(lapply(site_splits, make_summary_one_site), fill = TRUE)
log("took %.1f min", as.numeric(difftime(Sys.time(), t0, units = "mins")))

summaries <- summaries[!is.na(siteSR_id)]
```

## Export

Write the summary table: one row per site per day, with a column for each
variable and window (e.g., `precip_mm_prev7`, `tmax_degC_prev30`).
[03 · Split data](03_split_data.ipynb) joins it to the matchups on site and
anchor date.

```{r export}
write_feather(as.data.frame(summaries), output_summary_path)
log("wrote %s: %s rows, %s cols", output_summary_path, nrow(summaries), ncol(summaries))
```
