---
title: "Regional Application at In Situ SDD Locations"
author: "B Steele, Colorado State University, ROSSyndicate"
date: "last updated Sep 25, 2026"
output: html_document
---

**Pipeline step 07 · R · run from the repository root**

- **Purpose:** Apply the production ensemble to every Landsat observation at in situ SDD locations, apply the training-domain and area-of-applicability checks, and validate on the superseded matchups.
- **Inputs:** `aquamatch_files/`: `sdd_quality_filter.parquet`, `siteSR_DSWE1_regional_band_filtered.parquet`, `base_with_splits.parquet`, `regional_sdd_not_closest.parquet`, `site_characteristics.feather`, `downloads/siteSR_DSWE1_LS7corr.parquet`; `xg_models/v3_production/` (fold models, `ensemble_mean_abs_shap.csv`).
- **Outputs:** `aquamatch_files/application/`: `application_rs_estimates.parquet`, `application_filter_steps.csv`, `superseded_validation.csv`, and cached fetches; `xg_models/v3_production/applicability/`: `aoa_parameters.json`, `aoa_reference.csv`.
- **Previous:** [06 · Test-set sensitivity](06_test_set_sensitivity.ipynb) · **Used by:** `asv2_manuscript/manuscript_tables_figures.Rmd`

This script applies the production 10-seed XGBoost ensemble
(`04_make_models.ipynb`) to every Landsat observation at every location in the
region with quality-controlled in situ Secchi disc depth (SDD) data. It asks
how many additional SDD observations remote sensing provides at those same
locations.

That number is not simply the siteSR row count. An estimate only counts if
every one of these conditions holds:

1. the scene passes the same siteSR QA used to build the training data
   (`01_filter_AquaMatch_Data.Rmd`: geometry and regional HUC4 scoping, scene
   metadata QA, and band RANSAC);
2. it comes from a Landsat mission represented in training;
3. it falls within the seasonal window of the training matchups (the central
   98% of their day of year), which excludes ice-season scenes the model has
   effectively never seen;
4. every model input (bands, LakeCat catchment metrics, gridMET antecedent
   weather) is available;
5. every feature falls within the range of the training data (the literal
   "training box");
6. the observation falls inside the model's area of applicability (AOA;
   Meyer & Pebesma 2021), a multivariate dissimilarity check against the
   training data that is weighted by each feature's contribution to the model;
   and
7. the estimate is physically plausible: no shallower than the shallowest SDD
   in the training data. Tree ensembles can return small negative values for
   very turbid water.

Counts are reported at the level of physical *locations* rather than
`siteSR_id`s or `MonitoringLocationIdentifier`s. Following
`01_filter_AquaMatch_Data.Rmd`, sites within 100 m of each other on the same
waterbody are one location. Otherwise, a station listed under both its WQP and
its NWIS identifier would be counted twice on both the in situ and remote
sensing sides.

The slow steps (site characteristic fetches, gridMET pulls, and prediction) are
cached to `aquamatch_files/application/`. Re-running loads from cache. Set
`FORCE_REFRESH_CACHE <- TRUE` to rebuild. Outputs read by
`manuscript_tables_figures.Rmd` are written to
`aquamatch_files/application/`.

Before counting estimates, the ensemble is assessed on the matchups
`02a_make_matches.Rmd` set aside because a closer-in-time image existed for
the same in situ sample (see "Model performance on superseded matchups"). These are
image–field pairs from the same filtered record the model was built from,
passed through the same checks as the application.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

Load packages and define the run constants: input paths, the production
model directory and its 10 ensemble seeds (601–610), the output and cache
directory, and the cache-refresh switch. The model-input functions (site
characteristics, gridMET weather, feature engineering) come from
`R/prepare_model_inputs.R`, which end users call to prepare their own data, so
the application below runs the same code.

```{r libraries, message=FALSE, warning=FALSE}
library(arrow)
library(tidyverse)
library(sf)
library(igraph)
library(xgboost)
library(jsonlite)
library(RANN)

# model-input assembly shared with end users (site characteristics, gridMET
# weather, feature engineering); see the file header
source("regional_clarity/R/prepare_model_inputs.R")

log <- function(...) cat(sprintf("[%s] ", format(Sys.time(), "%H:%M:%S")), sprintf(...), "\n")

AQUAMATCH_DIR <- "regional_clarity/aquamatch_files"
SITES_META_PATH <- file.path(AQUAMATCH_DIR, "downloads", "siteSR_collated_WQP_NWIS_sites_with_NHD_info_2025-06-04.csv")
SITESR_RAW_PATH <- file.path(AQUAMATCH_DIR, "downloads", "siteSR_DSWE1_LS7corr.parquet")
SITESR_QA_PATH <- file.path(AQUAMATCH_DIR, "siteSR_DSWE1_regional_band_filtered.parquet")
SDD_QC_PATH <- file.path(AQUAMATCH_DIR, "sdd_quality_filter.parquet")
TRAINING_PATH <- file.path(AQUAMATCH_DIR, "base_with_splits.parquet")
MODEL_DIR <- file.path("regional_clarity", "xg_models", "v3_production")
ENSEMBLE_SEEDS <- 601:610

OUT_DIR <- file.path(AQUAMATCH_DIR, "application")
dir.create(OUT_DIR, showWarnings = FALSE)
FORCE_REFRESH_CACHE <- FALSE
N_WORKERS <- 8
```

## In situ SDD stations

The in situ baseline is the quality-controlled regional SDD record that feeds
matchup creation (`01_filter_AquaMatch_Data.Rmd`, the "RANSAC inliers" step):
tier/flag QC, the regional HUC4 restriction, and removal of per-site RANSAC
outliers. Sites with too few observations to run RANSAC (`is_inlier` missing)
are kept, as they are in `01`.

```{r insitu}
sdd <- read_parquet(SDD_QC_PATH,
                    col_select = c("MonitoringLocationIdentifier", "ActivityStartDate",
                                   "harmonized_value", "is_inlier")) %>%
  filter(is.na(is_inlier) | is_inlier) %>%
  mutate(date = as.Date(ActivityStartDate))

log("in situ SDD: %s rows, %s stations, %s station-days", nrow(sdd),
    n_distinct(sdd$MonitoringLocationIdentifier),
    nrow(distinct(sdd, MonitoringLocationIdentifier, date)))
```

## Link stations to siteSR sites and physical locations

AquaMatch's site table carries both the WQP-style
`MonitoringLocationIdentifier` (as `loc_id`) and, for NWIS-sourced sites, the
bare USGS station number (the WQP identifier is `USGS-` + that number). A
station can map to more than one `siteSR_id` (e.g. `WQP_*` and `NWIS_*`
entries for the same USGS station).

```{r link-sites}
sites_meta <- read_csv(SITES_META_PATH, show_col_types = FALSE,
                       col_select = c(siteSR_id, loc_id, source, WGS84_Latitude, WGS84_Longitude,
                                      wb_nhd_id, wb_nhd_source, flag_wb, flag_optical_shoreline)) %>%
  mutate(MonitoringLocationIdentifier = if_else(source == "NWIS", paste0("USGS-", loc_id), loc_id))

station_sites <- sites_meta %>%
  filter(MonitoringLocationIdentifier %in% sdd$MonitoringLocationIdentifier) %>%
  distinct(siteSR_id, .keep_all = TRUE)

stopifnot("some in situ stations have no AquaMatch site record" =
            all(unique(sdd$MonitoringLocationIdentifier) %in% station_sites$MonitoringLocationIdentifier))
log("%s stations -> %s siteSR_ids", n_distinct(station_sites$MonitoringLocationIdentifier),
    nrow(station_sites))
```

Group those `siteSR_id`s into physical locations with the same rule as
`01_filter_AquaMatch_Data.Rmd`: within 100 m and on the same NHD waterbody,
with grouping transitive through connected components. A site with no matched
waterbody is only grouped with itself.

```{r locations}
site_pts <- st_as_sf(station_sites, coords = c("WGS84_Longitude", "WGS84_Latitude"),
                     crs = 4326, remove = FALSE)
within_100m <- st_is_within_distance(site_pts, dist = 100)
wb <- station_sites$wb_nhd_id
within_100m_same_wb <- Map(function(nbrs, i) {
  keep <- nbrs == i | (!is.na(wb[nbrs]) & !is.na(wb[i]) & wb[nbrs] == wb[i])
  nbrs[keep]
}, within_100m, seq_along(within_100m))

station_sites <- station_sites %>%
  mutate(location_id = components(graph_from_adj_list(within_100m_same_wb, mode = "all"))$membership)

# one location per station: a station whose siteSR_ids land in different
# locations (rare; different coordinates across sources) is assigned to the
# location holding most of its siteSR_ids
station_location <- station_sites %>%
  count(MonitoringLocationIdentifier, location_id) %>%
  slice_max(n, n = 1, with_ties = FALSE, by = MonitoringLocationIdentifier) %>%
  select(MonitoringLocationIdentifier, location_id)

insitu_location_days <- sdd %>%
  inner_join(station_location, by = "MonitoringLocationIdentifier") %>%
  distinct(location_id, date)

log("%s stations -> %s physical locations; %s in situ location-days",
    nrow(station_location), n_distinct(station_location$location_id), nrow(insitu_location_days))
```

## siteSR observations at in situ locations

Step 0 is every siteSR observation at these locations before any QA, counted
from the national concatenated siteSR file. Step 1 is the QA'd regional
siteSR file written by `01_filter_AquaMatch_Data.Rmd` (geometry and HUC4
scoping, scene metadata QA, and band RANSAC). Because `01` scopes by HUC4 and
site geometry, a location can drop out entirely at step 1; those locations
are not observable in this framework.

```{r sitesr-rows}
location_lookup <- station_sites %>% select(siteSR_id, location_id)

raw_counts_cache <- file.path(OUT_DIR, "sitesr_raw_location_days.parquet")
if (!FORCE_REFRESH_CACHE && file.exists(raw_counts_cache)) {
  sitesr_raw_location_days <- read_parquet(raw_counts_cache)
} else {
  sitesr_raw_location_days <- open_dataset(SITESR_RAW_PATH) %>%
    select(siteSR_id, date) %>%
    filter(siteSR_id %in% location_lookup$siteSR_id) %>%
    collect() %>%
    mutate(date = as.Date(date)) %>%
    inner_join(location_lookup, by = "siteSR_id") %>%
    distinct(location_id, date)
  write_parquet(sitesr_raw_location_days, raw_counts_cache)
}

sitesr_qa <- read_parquet(SITESR_QA_PATH) %>%
  filter(siteSR_id %in% location_lookup$siteSR_id) %>%
  mutate(date = as.Date(date)) %>%
  inner_join(location_lookup, by = "siteSR_id")

log("pre-QA: %s location-days at %s locations", nrow(sitesr_raw_location_days),
    n_distinct(sitesr_raw_location_days$location_id))
log("post-QA: %s rows, %s location-days at %s locations", nrow(sitesr_qa),
    nrow(distinct(sitesr_qa, location_id, date)), n_distinct(sitesr_qa$location_id))
```

## Site characteristics

Elevation and LakeCat catchment metrics per `siteSR_id`, following
`02b_pull_site_characteristics.Rmd`. Sites already in that script's output
(`site_characteristics.feather`, the training sites) are reused, and only the
remaining sites are fetched. `shore_flag` uses the definition from
`03_split_data.ipynb`.

```{r site-characteristics}
site_char_cache <- file.path(OUT_DIR, "application_site_characteristics.feather")
qa_sites <- station_sites %>% filter(siteSR_id %in% unique(sitesr_qa$siteSR_id))

if (!FORCE_REFRESH_CACHE && file.exists(site_char_cache)) {
  site_char <- read_feather(site_char_cache)
} else {
  site_char <- fetch_site_characteristics(
    qa_sites, existing = read_feather(file.path(AQUAMATCH_DIR, "site_characteristics.feather")))
  write_feather(site_char, site_char_cache)
}

log("sites with LakeCat data: %s / %s", sum(!is.na(site_char$pct_impervious_2006)), nrow(site_char))
```

## Antecedent weather (gridMET)

gridMET is a 1/24° grid, and many sites share a cell. Weather is therefore
fetched once per cell (at the cell center) over the full date range needed by
that cell's siteSR observations, then assigned back to every site in the cell.
Rolling summaries follow `02c_pull_weather_summaries.Rmd` and cover the *w* days
strictly before the image date.

```{r gridmet-cells}
site_cells <- gridmet_cells(qa_sites)
cell_ranges <- gridmet_cell_ranges(sitesr_qa, site_cells)

log("%s sites fall in %s gridMET cells", nrow(site_cells), nrow(cell_ranges))
```

Fetch daily gridMET weather at each cell center in parallel (PSOCK workers,
see `02c_pull_weather_summaries.Rmd`). A failed cell is retried once, and cells
that still fail are written to `gridmet_fetch_failures.csv`. The raw daily
series are cached.

```{r gridmet-fetch}
weather_raw_cache <- file.path(OUT_DIR, "application_weather_raw.feather")

if (!FORCE_REFRESH_CACHE && file.exists(weather_raw_cache)) {
  weather_raw <- read_feather(weather_raw_cache)
} else {
  fetched <- fetch_gridmet(cell_ranges, n_workers = N_WORKERS)
  if (nrow(fetched$failed) > 0) write_csv(fetched$failed, file.path(OUT_DIR, "gridmet_fetch_failures.csv"))
  weather_raw <- fetched$weather
  write_feather(weather_raw, weather_raw_cache)
}
```

Convert units and pad each cell's series to one row per calendar day. Then
compute the 1-, 3-, 7-, and 30-day rolling summaries: total precipitation,
maximum daily high, mean temperature, minimum daily low, and mean solar
radiation. Dates are shifted by one day so that each window ends the day
before the image date.

```{r gridmet-summarize}
weather_summary_cache <- file.path(OUT_DIR, "application_weather_summaries.feather")

if (!FORCE_REFRESH_CACHE && file.exists(weather_summary_cache)) {
  weather_summaries <- read_feather(weather_summary_cache)
} else {
  weather_summaries <- summarize_weather(weather_raw)
  write_feather(as.data.frame(weather_summaries), weather_summary_cache)
}
log("weather summaries: %s cell-days", nrow(weather_summaries))
```

## Assemble model inputs

The band columns in the QA'd siteSR file are already harmonized to Landsat 7
(Gardner et al.; `*_corr_7`). Spectral indices and the coarsening of site
features are R ports of `python/features.py` (`add_spectral_indices()` and
`coarsen_site_features()`). The same function is applied to the training data
below, so the applicability checks compare like with like. Both are defined
in `R/prepare_model_inputs.R` (`add_features()`).

Build one model-input row per QA'd siteSR observation. Each row combines
the harmonized bands, a flag for Landsat 8/9 atmospheric correction
(LaSRC), site characteristics, and the antecedent weather for that site's
gridMET cell on the image date. Spectral indices and coarsened site
features are then added.

```{r assemble}
model_input <- assemble_model_inputs(sitesr_qa, site_char, site_cells, weather_summaries,
                                     keep = "location_id")

log("model input: %s rows", nrow(model_input))
```

## Predict with the 10-seed ensemble

Each seed's saved feature set and four fold models are used as trained. The
prediction for a row is the mean over all 40 models, and the spread across
the 10 seed means is kept as a measure of ensemble disagreement.

First, read the feature list each seed selected during tuning. Seeds can
differ slightly in their final features.

```{r model-features}
seed_feats <- map(ENSEMBLE_SEEDS, \(s) {
  fromJSON(file.path(MODEL_DIR, paste0("seed", s), "final_tune_xgboost.json"))$features
}) %>%
  set_names(ENSEMBLE_SEEDS)
model_feats <- sort(unique(unlist(seed_feats)))
```

For each seed, predict with its four fold models on that seed's features and
average them to give the seed's prediction. Across seeds, record the mean
(`pred_sdd`) and the standard deviation (`pred_seed_sd`). Predictions are
cached.

```{r predict}
pred_cache <- file.path(OUT_DIR, "application_predictions.parquet")

if (!FORCE_REFRESH_CACHE && file.exists(pred_cache)) {
  preds <- read_parquet(pred_cache)
} else {
  seed_preds <- map(ENSEMBLE_SEEDS, \(s) {
    feats <- seed_feats[[as.character(s)]]
    X <- as.matrix(model_input[, feats])
    fold_preds <- map(1:4, \(fold) {
      booster <- xgb.load(file.path(MODEL_DIR, paste0("seed", s), paste0("xgboost_fold", fold, ".json")))
      predict(booster, X)
    })
    log("seed%s predicted", s)
    rowMeans(do.call(cbind, fold_preds))
  })
  seed_mat <- do.call(cbind, seed_preds)
  preds <- tibble(pred_sdd = rowMeans(seed_mat), pred_seed_sd = apply(seed_mat, 1, sd))
  write_parquet(preds, pred_cache)
}
stopifnot(nrow(preds) == nrow(model_input))
model_input <- bind_cols(model_input, preds)
```

## Applicability checks

### Training reference

The reference is the cross-validation pool that trained the ensemble (every
non-holdout row of `base_with_splits.parquet`), with features engineered by
the same function as above.

```{r training-reference}
training <- read_parquet(TRAINING_PATH) %>%
  mutate(date = as.Date(date)) %>%
  add_features()
train_pool <- training %>% filter(!is_holdout)
log("training reference: %s CV-pool rows", nrow(train_pool))
```

### Feature range ("training box")

An observation passes when every model input used by any seed lies within the
minimum and maximum of that feature in the training pool. Tree ensembles do
not extrapolate: outside the training range, a prediction flattens at the
value of the outermost leaf, so these estimates carry no information about
how SDD responds beyond the sampled conditions.

```{r range-check}
feature_ranges <- train_pool %>%
  summarise(across(all_of(model_feats), list(min = ~ min(.x, na.rm = TRUE), max = ~ max(.x, na.rm = TRUE)))) %>%
  pivot_longer(everything(), names_to = c("feature", ".value"), names_pattern = "(.*)_(min|max)$")

in_range <- function(df) {
  ok <- rep(TRUE, nrow(df))
  for (i in seq_len(nrow(feature_ranges))) {
    x <- df[[feature_ranges$feature[i]]]
    ok <- ok & !is.na(x) & x >= feature_ranges$min[i] & x <= feature_ranges$max[i]
  }
  ok
}

feature_complete <- function(df) complete.cases(df[, model_feats])
```

### Seasonal window

The training matchups are concentrated in the open-water season (about 85% in
May–October; under 4% in November–March). The area of applicability below
weights features by their SHAP contribution, and air and surface temperature
carry little weight, so it barely distinguishes a January scene from a July
scene with similar optics. Ice-season conditions (partial ice cover, winter
mixing, snowmelt) are nonetheless outside what the model has learned. An
observation therefore has to fall within the day-of-year window spanning the
central 98% of the training matchups (1st to 99th percentile).

```{r season-window}
training_doy <- yday(train_pool$date)
season_window <- round(quantile(training_doy, c(0.01, 0.99)))
in_season <- function(df) between(yday(df$date), season_window[1], season_window[2])
log("seasonal window: day of year %s-%s", season_window[1], season_window[2])
```

### Area of applicability

The dissimilarity index (DI) of Meyer & Pebesma (2021) is computed over the
features every seed kept (the same core used for SHAP attribution in
`05_evaluate_ensemble.ipynb`). Features are standardized by the training
pool's mean and standard deviation, then weighted by their ensemble mean
|SHAP|, so distance is measured in the directions the model relies on. An
observation's DI is the distance to its nearest training observation, divided
by the mean pairwise distance within the training pool.

The threshold follows Meyer & Pebesma (2021): the DI of each training
observation to its nearest neighbor *in a different cross-validation fold*
(the four HUC8-grouped CV-pool partitions, `holdout_part` 1-4), with the
threshold set at the upper whisker of that distribution (third quartile +
1.5 × IQR).

```{r aoa}
shap_weights <- read_csv(file.path(MODEL_DIR, "ensemble_mean_abs_shap.csv"), show_col_types = FALSE)
aoa_feats <- shap_weights$feature
stopifnot(all(aoa_feats %in% model_feats))

centers <- map_dbl(aoa_feats, \(f) mean(train_pool[[f]], na.rm = TRUE))
scales <- map_dbl(aoa_feats, \(f) sd(train_pool[[f]], na.rm = TRUE))
weights <- shap_weights$mean_abs_shap

to_aoa_space <- function(df) {
  X <- as.matrix(df[, aoa_feats])
  X <- sweep(sweep(X, 2, centers), 2, scales, "/")
  sweep(X, 2, weights, "*")
}

# the handful of training rows missing a weather summary (~0.1%) can't be
# placed in feature space and are left out of the reference
train_X <- to_aoa_space(train_pool)
train_complete <- complete.cases(train_X)
train_X <- train_X[train_complete, ]
log("AOA reference: %s of %s CV-pool rows complete", nrow(train_X), length(train_complete))

# mean pairwise distance within the training pool
d_bar <- mean(dist(train_X))

# cross-fold nearest-neighbor DI for the threshold
train_fold <- train_pool$holdout_part[train_complete]
train_di <- numeric(nrow(train_X))
for (k in unique(train_fold)) {
  in_k <- train_fold == k
  train_di[in_k] <- nn2(train_X[!in_k, ], train_X[in_k, ], k = 1)$nn.dists[, 1] / d_bar
}
aoa_threshold <- boxplot.stats(train_di)$stats[5]

compute_di <- function(df) {
  di <- rep(NA_real_, nrow(df))
  X <- to_aoa_space(df)
  ok <- complete.cases(X)
  di[ok] <- nn2(train_X, X[ok, , drop = FALSE], k = 1)$nn.dists[, 1] / d_bar
  di
}

log("AOA threshold (DI): %.3f", aoa_threshold)
```

## Apply the filters

Each step is counted in location-days: the number of distinct
(location, date) pairs with at least one observation remaining. When more than
one scene passes at a location on the same day (overlapping paths, or several
`siteSR_id`s at one location), the day's estimate is their mean.

```{r filter-steps}
model_input <- model_input %>%
  mutate(pass_mission = mission %in% TRAINING_MISSIONS,
         pass_season = pass_mission & in_season(.),
         pass_complete = pass_season & feature_complete(.),
         pass_range = pass_complete & in_range(.),
         di = compute_di(.),
         pass_aoa = pass_range & !is.na(di) & di <= aoa_threshold,
         pass_plausible = pass_aoa & pred_sdd >= min(train_pool$harmonized_value))

count_step <- function(df, step) {
  tibble(step = step, location_days = nrow(distinct(df, location_id, date)),
         locations = n_distinct(df$location_id))
}

filter_steps <- bind_rows(
  count_step(sitesr_raw_location_days, "siteSR observations (before QA)"),
  count_step(model_input, "After siteSR QA"),
  count_step(filter(model_input, pass_mission), "Landsat mission represented in training"),
  count_step(filter(model_input, pass_season), "Within seasonal window of training matchups"),
  count_step(filter(model_input, pass_complete), "All model inputs available"),
  count_step(filter(model_input, pass_range), "Within training range of every input"),
  count_step(filter(model_input, pass_aoa), "Inside area of applicability"),
  count_step(filter(model_input, pass_plausible), "Estimate no shallower than training minimum")
)
filter_steps

rs_estimates <- model_input %>%
  filter(pass_plausible) %>%
  summarise(pred_sdd = mean(pred_sdd),
            pred_seed_sd = mean(pred_seed_sd),
            di = max(di),
            n_scenes = n(),
            .by = c(location_id, date))
```

## Export

Write the application outputs:

- the step-by-step filter counts;
- the per-location-day remote sensing estimates;
- the in situ location-days for comparison;
- location coordinates;
- per-row pass/fail flags for every check.

The applicability checks themselves are written alongside the model, to
`xg_models/v3_production/applicability/`, so anyone applying the model can
screen their own observations without the training data:

- `aoa_parameters.json`: the AOA threshold and mean training distance; the
  AOA features with their standardization centers, scales, and |SHAP|
  weights; the training range of every model input; the seasonal window; the
  training missions; and the training minimum SDD;
- `aoa_reference.csv`: the training reference already transformed into AOA
  space (standardized and weighted), one row per CV-pool matchup, which is
  what a new observation's nearest-neighbor distance is measured against.

```{r export}
locations <- station_sites %>%
  summarise(lat = mean(WGS84_Latitude), lon = mean(WGS84_Longitude), .by = location_id)

write_csv(filter_steps, file.path(OUT_DIR, "application_filter_steps.csv"))
write_parquet(rs_estimates, file.path(OUT_DIR, "application_rs_estimates.parquet"))
write_parquet(insitu_location_days, file.path(OUT_DIR, "application_insitu_location_days.parquet"))
write_parquet(locations, file.path(OUT_DIR, "application_locations.parquet"))
write_parquet(model_input %>%
                select(location_id, date, mission, pass_mission, pass_season, pass_complete, pass_range, pass_aoa, pass_plausible, di),
              file.path(OUT_DIR, "application_row_flags.parquet"))
log("wrote application outputs to %s", OUT_DIR)

APPLICABILITY_DIR <- file.path(MODEL_DIR, "applicability")
dir.create(APPLICABILITY_DIR, showWarnings = FALSE)
write_json(list(aoa_threshold = aoa_threshold,
                d_bar = d_bar,
                aoa_features = tibble(feature = aoa_feats, center = centers,
                                      scale = scales, weight = weights),
                feature_ranges = feature_ranges,
                season_window_doy = unname(season_window),
                training_missions = TRAINING_MISSIONS,
                min_training_sdd_m = min(train_pool$harmonized_value)),
           file.path(APPLICABILITY_DIR, "aoa_parameters.json"),
           auto_unbox = TRUE, digits = I(17), pretty = TRUE)
write_csv(as_tibble(train_X), file.path(APPLICABILITY_DIR, "aoa_reference.csv"))
log("wrote applicability checks to %s", APPLICABILITY_DIR)
```

## Model performance on superseded matchups

This is a true side-quest: `02a_make_matches.Rmd` pairs each in situ SDD sample with every rain-screened
siteSR observation within ±5 days, keeps the one closest in time for
training, and sets the rest aside (`regional_sdd_not_closest.parquet`). Those
superseded matchups are real image–field pairs that passed every upstream
filter but never entered model training as rows, which makes them a direct
test of the application setting: an image reflectance sumamry that is not the one the model was
trained on, at a monitored location. Each in situ sample behind them also has
a closest-in-time match in the training data, so the field value itself is
not new to the model at sites in the cross-validation pool, rather the reflectance
summaries are new to the model. The performance of the model across these
data represent something between a true hold out spatially and members of 
other cross-validation groups - so it should land somewhere between the 
cross-validation statisitcs and the test set statistics. 

Features follow `03_split_data.ipynb` exactly: harmonized bands from the
matchup, the same site characteristics, and antecedent weather anchored on the
later of the image and field dates. The 40-model ensemble is applied as
above, and the same applicability checks and plausibility floor are applied
as in the regional application.

```{r superseded-performance}
superseded <- read_parquet(file.path(AQUAMATCH_DIR, "regional_sdd_not_closest.parquet")) %>%
  transmute(siteSR_id, date = as.Date(date), mission, obs = harmonized_value,
            field_date = as.Date(harmonized_local_time, tz = "UTC"),
            anchor_date = pmax(date, field_date),
            red_corr7 = red_corr_7, green_corr7 = green_corr_7, blue_corr7 = blue_corr_7,
            nir_corr7 = nir_corr_7, swir1_corr7 = swir1_corr_7, swir2_corr7 = swir2_corr_7,
            temp_corr7 = surfacetemp_corr_7,
            atm_corr_LaSRC = as.integer(mission %in% c("LC08", "LC09"))) %>%
  left_join(site_char, by = "siteSR_id") %>%
  left_join(site_cells %>% select(siteSR_id, cell_id), by = "siteSR_id") %>%
  left_join(as.data.frame(weather_summaries), by = c("cell_id", "anchor_date" = "date")) %>%
  add_features() %>%
  as_tibble()

superseded$pred <- map(ENSEMBLE_SEEDS, \(s) {
  X <- as.matrix(superseded[, seed_feats[[as.character(s)]]])
  map(1:4, \(fold) {
    predict(xgb.load(file.path(MODEL_DIR, paste0("seed", s), paste0("xgboost_fold", fold, ".json"))), X)
  })
}) %>%
  list_flatten() %>%
  do.call(cbind, .) %>%
  rowMeans()

superseded <- superseded %>%
  left_join(training %>% distinct(siteSR_id, is_holdout), by = "siteSR_id") %>%
  mutate(pass_mission = mission %in% TRAINING_MISSIONS,
         pass_season = pass_mission & in_season(.),
         pass_complete = pass_season & feature_complete(.),
         pass_range = pass_complete & in_range(.),
         di = compute_di(.),
         pass_aoa = pass_range & !is.na(di) & di <= aoa_threshold,
         pass_plausible = pass_aoa & pred >= min(train_pool$harmonized_value))

log("superseded matchups: %s rows at %s sites (%s at holdout sites); %s pass every check",
    nrow(superseded), n_distinct(superseded$siteSR_id), sum(superseded$is_holdout, na.rm = TRUE),
    sum(superseded$pass_plausible))

superseded_metrics <- function(df) {
  summarise(df, n = n(),
            rmse = sqrt(mean((pred - obs)^2)),
            mae = mean(abs(pred - obs)),
            bias = mean(pred - obs),
            r2 = cor(pred, obs)^2)
}

superseded_validation <- bind_rows(
  superseded %>%
    mutate(group = case_when(!pass_range ~ "outside season or training range",
                             !pass_aoa ~ "outside AOA",
                             !pass_plausible ~ "below plausibility floor",
                             TRUE ~ "passes all checks")),
  superseded %>% mutate(group = "all superseded")
) %>%
  mutate(sites = if_else(is_holdout, "holdout sites", "CV-pool sites")) %>%
  bind_rows(mutate(., sites = "all sites")) %>%
  group_by(sites, group) %>%
  superseded_metrics() %>%
  ungroup()

write_csv(superseded_validation, file.path(OUT_DIR, "superseded_validation.csv"))
write_parquet(superseded %>% select(siteSR_id, date, field_date, mission, obs, pred, is_holdout, di, starts_with("pass_")),
              file.path(OUT_DIR, "superseded_predictions.parquet"))
superseded_validation %>% mutate(across(c(rmse, mae, bias, r2), ~ round(.x, 3)))
```

Plot ensemble estimates against in situ SDD for the superseded matchups that
pass every check. Panels show all sites, holdout sites, and
cross-validation-pool sites, each annotated with its error metrics. The
dashed line is 1:1.

```{r superseded-plot, fig.width=10, fig.height=4.5}
superseded_passing <- superseded %>%
  filter(pass_plausible) %>%
  mutate(sites = if_else(is_holdout, "holdout sites", "CV-pool sites")) %>%
  bind_rows(mutate(., sites = "all sites")) %>%
  mutate(sites = factor(sites, levels = c("all sites", "holdout sites", "CV-pool sites")))

panel_labels <- superseded_validation %>%
  filter(group == "passes all checks") %>%
  mutate(sites = factor(sites, levels = levels(superseded_passing$sites)),
         label = sprintf("n = %s\nRMSE = %.2f m\nMAE = %.2f m\nbias = %+.2f m\nR² = %.2f",
                         format(n, big.mark = ",", trim = TRUE), rmse, mae, bias, r2))

axis_max <- ceiling(max(superseded_passing$obs, superseded_passing$pred))

ggplot(superseded_passing, aes(obs, pred)) +
  geom_abline(linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.3, size = 1) +
  geom_text(data = panel_labels, aes(x = 0, y = axis_max, label = label),
            hjust = 0, vjust = 1, size = 3.2, lineheight = 0.9) +
  coord_equal(xlim = c(0, axis_max), ylim = c(0, axis_max)) +
  facet_wrap(~ sites) +
  labs(x = "in situ SDD (m)", y = "ensemble estimate (m)",
       title = "Superseded matchups that pass every applicability check") +
  theme_bw()
```

